---
title: "Creating species-level dataset (Level 2)"
author: "Patrick Pata"
date: "01/12/2022"
output: html_document
---

This file loads the level 1 dataset and calculates the species-level averages for numerical traits, merges the sources of repeated categorical trait records, and adds the binary version of categorical traits. 

The maximum body length trait is calculated for each species. The female or adult total length is selected when possible and the average size is calculated resulting in the mean maximum body length. If a value is not found for a species, the maximum length value across life stages and observations is selected. 

For all other numerical traits, the species level average is calculated. For categorical traits that have similar trait values but have different references are merged into a single row with multiple references for the same observation. Categorical traits are then converted into a binary format. Both the categorical and binary versions are retained in the level 2 trait dataset. Species-level trait values are then derived from numerical traits when a related trait has records.

*NOTE* This code demonstrates how the level 2 data is generated. This will not export the data file. The function updateIDs() is commented here as to not generate new ID numbers and mess with the master_id_list file since the level 2 data file is already generated.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Load libraries and functions
```{r}
packages <- c("tidyverse", "lubridate")

# Function to download the packages if necessary. Otherwise, these are loaded.
package.check <- lapply(
  packages,
  FUN = function(x)
  {
    if (!require(x, character.only = TRUE))
    {
      install.packages(x, dependencies = TRUE,
                       repos = "http://cran.us.r-project.org")
      library(x, character.only = TRUE)
    }
  }
)

source("toolkit.R")
```

# Prepare files and folders
```{r}
s.format <- read.csv("data_input/trait_dataset_standard_format_20230628.csv")[-1,]

trait.directory <- read.csv("data_input/trait_directory_20230628.csv") %>% 
  distinct(traitID, .keep_all = TRUE)

# taxonomy table
taxonomy <- read.csv("data_input/taxonomy_table_20230628.csv") %>% 
  distinct(taxonID, .keep_all = TRUE)

# stage table
lifestagelist <- read.csv("data_input/lifestage_directory_20230628.csv") %>% 
  select(-c(majorgroup, notes))


infol <- "data_input/Trait_dataset_level1/"
outfol <- "data_input/Trait_dataset_level2/"

# Load Level 1 dataset
traits.lvl1 <- read.csv(paste0(infol,"trait_dataset_level1-2023-08-15.csv"))
```

# Calculate bodyLengthMax trait
```{r}
trait.num <- traits.lvl1 %>% 
  filter(valueType == "numeric") %>% 
  mutate(traitValue = as.numeric(traitValue))

A <- trait.num %>% 
  filter(traitName == "bodyLength")

# For copepods, select females and total length when possible
A.1 <- A %>% 
  filter(class == "Copepoda") %>% 
  filter(lifeStage == "female" & sizeType == "total length")
# females but not in total length
A.2 <- A %>% 
  filter(class == "Copepoda") %>% 
  filter(taxonID %notin% A.1$taxonID) %>% 
  filter(lifeStage == "female")

# then select "adult"
A.3 <- A %>% 
  filter(taxonID %notin% A.1$taxonID) %>% 
  filter(taxonID %notin% A.2$taxonID) %>% 
  filter(lifeStage %in% c("adult"))

# Get average for adults and females
A.4 <- bind_rows(A.1, A.2, A.3) 

# all taxa not in A.4
B <- A %>% 
  filter(taxonID %notin% A.4$taxonID) 

# Update IDs, all are new observations because the trait is introduced here.
trait.length <- bind_rows(A.4, B) %>% 
  group_by(taxonID) %>% 
  mutate(individualCount = n(),
         dispersionSD = sd(traitValue, na.rm = TRUE),
         traitValue = mean(traitValue, na.rm = TRUE),
         notes = paste0("taxon level average; ",notes),
         # mark as derived
         isDerived = TRUE,
         catalogSource = paste(catalogNumber, collapse = "; ")) %>% 
  mergeTraitDetails(trait.directory) %>%
  mutate(traitName = "bodyLengthMax",
         basisOfRecord = "calculated taxon average",
         observationNumber = 1, maxObsNum = 1,
         catalogNumber = NA) %>% 
  standardizeID(trait.directory)
  # updateIDs()

rm(A, A.1, A.2, A.3, A.4, B)
```

# Calculate species-level averages for numerical traits
```{r}
# Calculate the mean associated size for each trait-taxon pair
sizeAssocSum <- summarise.sizeAssoc(trait.num)

trait.ave <- trait.num %>% 
  filter(traitName != "bodyLength") %>% 
  # exclude values not at standard temperature of 15 deg C for respiration, excretion, feeding, and growth rates
  filter(!(traitID %in% c(75:137) & assocTemperature != 15)) %>% 
  # if N is NA or if sd is not reported, N = 1
  mutate(individualCount = if_else(is.na(individualCount), 1, individualCount)) %>% 
  mutate(individualCount = if_else(is.na(dispersionSD), 1, individualCount)) %>% 
  # calculate trait-taxon average
  group_by(traitID, taxonID, traitUnit) %>%  
  # TODO: fix this mutate so it doesn't do the calculation thrice
  mutate(nRecords = n(),
         tvN = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$N,
         tvSD = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$sd,
         tvVal = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$mean,
         traitValue = tvVal, individualCount = tvN, dispersionSD = tvSD,
         notes = paste0("taxon level average; ",notes),
         aggregateMeasure = TRUE,
         basisOfRecord = "calculated taxon average") %>% 
  dplyr::select(-c(tvVal, tvN, tvSD)) %>% 
  mergeTraitDetails(trait.directory) %>% 
  ungroup() %>% 
  # update the size assoc columns
  dplyr::select(-c(sizeAssocName, sizeAssocValue, 
                   sizeAssocSD, sizeAssocN,
                   sizeAssocUnit, sizeAssocReference)) %>% 
  left_join(sizeAssocSum, by = c("traitID","taxonID")) 

# Filter which records are based on more than 1 observation and merge with trait.num.
trait.ave.A <- trait.ave %>% 
  filter(nRecords == 1)
trait.ave.B <- trait.ave %>% 
  filter(nRecords > 1) %>% 
  select(-nRecords)
  # remove catalogNumber before merging
  # mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA) %>% 
  # updateIDs()
  
trait.num <- trait.num %>% 
  filter(catalogNumber %in% trait.ave.A$catalogNumber) %>% 
  # merge the species averages
  bind_rows(trait.ave.B) %>% 
  # merge bodyLengthMax trait
  bind_rows(trait.length)

rm(trait.ave.A, trait.ave.B, trait.length, trait.ave)
```

# Merge references of categorical traits with similar values 
```{r}
# Merge records with similar values
trait.cat <- traits.lvl1 %>% 
  filter(valueType == "categorical") %>% 
  group_by(traitID, taxonID, stageID, traitValue, traitUnit) %>% 
  mergeTraitDetails(trait.directory)

# Records not merged
trait.cat.A <- trait.cat %>% 
  filter(is.na(catalogSource))
# Merged records
trait.cat.B <- trait.cat %>% 
  filter(!is.na(catalogSource)) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA)

# Regenerate the categorical traits df with the original records + merged records
trait.cat <- traits.lvl1 %>% 
  filter(catalogNumber %in% trait.cat.A$catalogNumber) %>% 
  bind_rows(trait.cat.B)
  # updateIDs()

```

# Convert categorical traits to binary
- Traits with binary matrices: feedingMode, trophicGroup, reproductiveMode, verticalDistribution, myelination, bioluminescence, dvm (+reverse), positive instances of hibernation at any stage
```{r}
# Trophic group
tlist <- c("herbivore", "omnivore", "carnivore", "detritivore", 
           "parasite", "commensal")
trait.TG <- trait.cat %>% 
  filter(traitName == "trophicGroup") %>% 
  mutate(TG.carnivore = if_else(grepl("carnivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.omnivore = if_else(grepl("omnivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.herbivore = if_else(grepl("herbivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.detritivore = if_else(grepl("detritivore",traitValue)==TRUE,1,0)) 
# Feeding mode
# Currently, passive == ambush (not considering commensal/parasite) and particle feeding
# Ambush feeding has separate column and can be classified into active/passive
tlist <- c("active","passive","cruise","current","active ambush","passive ambush",
           "ambush","particle feeder","commensal","parasite")
trait.FM <- trait.cat %>% 
  filter(traitName == "feedingMode") %>% 
  mutate(FM.active = if_else(grepl("active",traitValue)==TRUE & 
                          grepl("active ambush",traitValue)==FALSE,
                          1,0)) %>%  
  mutate(FM.passive = if_else(grepl("passive",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.cruise = if_else(grepl("cruise",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.current = if_else(grepl("current",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.active.ambush = if_else(grepl("active ambush",
                                       traitValue)==TRUE,1,0)) %>% 
  mutate(FM.passive.ambush = if_else(grepl("passive ambush",
                                        traitValue)==TRUE,1,0)) %>% 
  mutate(FM.ambush = if_else(grepl("ambush",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.particle.feeder = if_else(grepl("particle feeder",
                                         traitValue)==TRUE,1,0)) %>% 
  mutate(FM.commensal = if_else(grepl("commensal",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.parasite = if_else(grepl("parasite",traitValue)==TRUE,1,0)) %>% 
  # assign current and cruise as active & ambush and particle feeder as passive 
  mutate(FM.active = if_else(FM.cruise == 1 | FM.current == 1,
                             1, FM.active)) %>% 
  mutate(FM.passive = if_else(FM.ambush == 1 | FM.particle.feeder == 1, 
                              1, FM.passive))

# Reproduction mode
tlist <- c("broadcast","brooding","asexual","polyp")
trait.RM <- trait.cat %>% 
  filter(traitName == "reproductionMode") %>% 
  mutate(RM.brooding = if_else(grepl("brooding",traitValue)==TRUE,1,0)) %>% 
  mutate(RM.broadcasting = if_else(grepl("broadcasting",traitValue)==TRUE,1,0)) %>% 
  # assign polyp as asexual
  mutate(RM.asexual = if_else(grepl("asexual",traitValue)==TRUE |
                              grepl("polyp", traitValue)==TRUE,1,0)) 

# Vertical distribution
tlist <- c("epi","meso","bathy")
trait.VD <- trait.cat %>% 
  filter(traitName == "verticalDistribution") %>% 
  mutate(VD.epipelagic = if_else(grepl("epi",traitValue)==TRUE,1,0)) %>% 
  mutate(VD.mesopelagic = if_else(grepl("meso",traitValue)==TRUE,1,0)) %>% 
  mutate(VD.bathypelagic = if_else(grepl("bathy",traitValue)==TRUE,1,0)) %>% 
  # if epibathypelagic, meso is true
  mutate(VD.mesopelagic = if_else(VD.epipelagic == 1 & VD.bathypelagic == 1,
                                  1, VD.mesopelagic))

# Myelination
tlist <- c("absent","present","myelin-like","likely absent","likely present")
trait.MYE <- trait.cat %>% 
  filter(traitName == "myelination") %>% 
  # consider myelin-like as true, exclude likely
  mutate(MYE.absent = if_else(traitValue == "absent",1,0),
         MYE.present = if_else(traitValue %in% c("present","myelin-like"),1,0)) %>% 
  filter(MYE.absent == 1 | MYE.present == 1)

# 8-Bioluminescence
tlist <- c("absent","present","uncertain")
trait.BIO <- trait.cat %>% 
  filter(traitName == "bioluminescence") %>% 
  mutate(BIO.absent = if_else(traitValue == "absent",1,0),
         BIO.present = if_else(traitValue == "present",1,0)) %>% 
  # disregard uncertain
  filter(BIO.absent == 1 | BIO.present == 1)

# DVM
tlist <- c("absent","present","reverse","weak","strong")
trait.DVM <- trait.cat %>% 
  filter(traitName == "dielVerticalMigration") %>% 
  mutate(DVM.absent = if_else(grepl("absent", traitValue)==TRUE,1,0),
         DVM.present = if_else(grepl("present", traitValue)==TRUE,1,0),
         DVM.reverse = if_else(grepl("reverse", traitValue)==TRUE,1,0),
         DVM.strong = if_else(grepl("strong", traitValue)==TRUE,1,0),
         DVM.weak = if_else(grepl("weak", traitValue)==TRUE,1,0)) %>% 
    # present here is only day at depth, night at surface DVM
  mutate(DVM.present = if_else(DVM.reverse == 1, 0, DVM.present))    
  
# Hibernation - only positive instances at any stage
# merge resting eggs with hibernation (hibernation at egg stage)
trait.HIB <- trait.cat %>% 
  filter(traitName %in% c("hibernationStage", "restingEggs")) %>% 
  mutate(HIB.present = 1)

# TODO assign the source traitIDs here, remove verbatim records

# Combine binary matrices
trait.binary <- bind_rows(trait.TG, trait.FM, trait.RM, trait.VD,
                          trait.MYE, trait.DVM, trait.HIB, trait.BIO) %>% 
  # Set as derived and assign the source traitIDs
  mutate(isDerived = TRUE,
         basisOfRecord = "converted from categorical",
         catalogSource = catalogNumber) %>% 
  # Remove verbatim columns
  removeVerbatimRecords() %>% 
  relocate(colnames(s.format))

# List the binary traitNames for pivoting
binaryTraitNames <- data.frame(names = colnames(trait.binary)) %>% 
  filter(grepl("TG|FM|RM|VD|MYE|DVM|HIB|BIO",names))
trait.binary <- trait.binary %>% 
  select(-c(traitName, traitValue)) %>% 
  pivot_longer(cols = all_of(binaryTraitNames$names), 
               names_to = "traitName", values_to = "traitValue") %>% 
  mutate(valueType = "binary", maxObsNum = 0) %>% 
  filter(traitValue == 1) %>%
  distinct(traitName, traitValue, taxonID, .keep_all = TRUE) %>% 
  # Assign trait numbers and update trait-taxon-ID
  standardizeID(trait.directory) %>% 
  # mutate(catalogNumber = NA) %>% 
  # updateIDs() %>% 
  mutate(traitValue = as.character(traitValue))

```


# Curate size: Identify instances when CW > DW > WW
```{r}
size.check <- trait.num %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(traitName %in% c("carbonWeight","dryWeight","wetWeight")) %>% 
  select(taxonID, scientificName, traitName, traitValue) %>% 
  pivot_wider(names_from = traitName, values_from = traitValue) %>% 
  mutate(cw.gr = if_else(carbonWeight > dryWeight, T, F),
         dw.gr = if_else(dryWeight > wetWeight, T, F))

# Exclude wet weight when dry weight is larger
size.check.ww <- size.check %>%
  filter(dw.gr == T)

trait.num <- trait.num %>% 
  filter(!(traitName == "wetWeight" & taxonID %in% size.check.ww$taxonID))

# Exclude dry weight when carbon weight is larger
size.check.dw <- size.check %>%
  filter(cw.gr == T)

trait.num <- trait.num %>% 
  filter(!(traitName == "dryWeight" & taxonID %in% size.check.dw$taxonID))
```


# Derive some species level trait values
## Water content
```{r}
# waterPWW
A <- trait.num %>% 
  filter(traitName == "waterPWW")

# Calculate water content as inverse of dry weight % wet weight
B <- trait.num %>% 
  filter(traitName  == "dryWeightPWW") %>% 
  filter(taxonID %notin% A$taxonID) %>% 
  group_by(traitID, taxonID) %>%
  # Derived columns are the original columns
  mutate(verbatimScientificName = scientificName,
         verbatimTraitName = traitName,
         verbatimTraitValue = as.character(traitValue),
         verbatimTraitUnit = traitUnit,
         verbatimLifeStage = lifeStage,
         verbatimNotes = notes,
         traitName = "waterPWW", traitID = 19,
         traitValue = 100 - traitValue,
         dispersionSD = NaN,
         individualCount = 1,
         basisOfRecord = "derived from related trait",
         aggregateMeasure = FALSE,
         isDerived = TRUE,
         catalogSource = catalogNumber,
         notes = paste0("Calculated from dryWeightPWW based on ", catalogNumber),
         maxObsNum = 0) %>% 
  ungroup() %>% 
  standardizeID(trait.directory) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA) %>% 
  mutate(catalogNumber = as.character(catalogNumber))

# waterPWW
C <- trait.num %>% 
  filter(traitName == "dryWeightPWW")

# Calculate water content as inverse of dry weight % wet weight
D <- trait.num %>% 
  filter(traitName  == "waterPWW") %>% 
  filter(taxonID %notin% A$taxonID) %>% 
  group_by(traitID, taxonID) %>%
  # Derived columns are the original columns
  mutate(verbatimScientificName = scientificName,
         verbatimTraitName = traitName,
         verbatimTraitValue = as.character(traitValue),
         verbatimTraitUnit = traitUnit,
         verbatimLifeStage = lifeStage,
         verbatimNotes = notes,
         traitName = "dryWeightPWW", traitID = 20,
         traitValue = 100 - traitValue,
         dispersionSD = NaN,
         individualCount = 1,
         basisOfRecord = "derived from related trait",
         aggregateMeasure = FALSE,
         isDerived = TRUE,
         catalogSource = catalogNumber,
         notes = paste0("Calculated from waterPWW based on ", catalogNumber),
         maxObsNum = 0) %>% 
  ungroup() %>% 
  standardizeID(trait.directory) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA) %>% 
  mutate(catalogNumber = as.character(catalogNumber))

# Add the derived water content to trait.num
trait.num <- trait.num %>% 
  bind_rows(B, D)
  # updateIDs()

rm(A,B,C,D)
```


## Estimate egg carbon content from egg diameter
```{r}
A <- trait.num %>% 
  filter(traitName == "eggWeight")

B <- trait.num %>% 
  filter(traitName  == "eggDiameter") %>% 
  filter(taxonID %notin% A$taxonID) %>% 
  filter(traitUnit == "um") %>% 
  group_by(traitID, taxonID) %>% 
  # Derived columns are the original columns
  mutate(verbatimScientificName = scientificName,
         verbatimTraitName = traitName,
         verbatimTraitValue = as.character(traitValue),
         verbatimTraitUnit = traitUnit,
         verbatimLifeStage = lifeStage,
         verbatimNotes = notes,
         traitName = "eggWeight", traitID = 153,
         traitValue = (4/3)*pi*(traitValue/2)^3 * (0.14*10^-6),
         traitUnit = "ug",
         dispersionSD = NaN,
         individualCount = 1,
         basisOfRecord = "derived from related trait",
         aggregateMeasure = FALSE,
         isDerived = TRUE,
         catalogSource = catalogNumber,
         notes = paste0("Estimated from egg diameter; ", catalogNumber),
         maxObsNum = 0) %>% 
  ungroup() %>% 
  standardizeID(trait.directory) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA)

# Add the derived water content to trait.num
trait.num <- trait.num %>% 
  bind_rows(B) 
  # updateIDs()

rm(A, B)
```


## Composition
```{r}
ww <- trait.num %>% 
  filter(traitName == "wetWeight") %>% 
  mutate(sizeAssocReference = if_else(!is.na(secondaryReference),
                                      secondaryReference, primaryReference)) %>% 
  select(taxonID, sizeAssocName = traitName, 
         sizeAssocValue = traitValue, sizeAssocUnit = traitUnit,
         sizeAssocReference)
dw <- trait.num %>% 
  filter(traitName == "dryWeight") %>% 
  mutate(sizeAssocReference = if_else(!is.na(secondaryReference),
                                      secondaryReference, primaryReference)) %>% 
  select(taxonID, sizeAssocName = traitName, 
         sizeAssocValue = traitValue, sizeAssocUnit = traitUnit,
         sizeAssocReference)
cw <- trait.num %>% 
  filter(traitName == "carbonWeight") %>% 
  mutate(sizeAssocReference = if_else(!is.na(secondaryReference),
                                      secondaryReference, primaryReference)) %>% 
  select(taxonID, sizeAssocName = traitName, 
         sizeAssocValue = traitValue, sizeAssocUnit = traitUnit,
         sizeAssocReference)

# Carbon
C1 <- convertTotal2PW(trait.num, dw, "carbonPDW", "carbonWeight")
C2 <- convertTotal2PW(trait.num, ww, "carbonPWW", "carbonWeight") # can contain some really large values
C3 <- convertPW2Total(trait.num, cw, "carbonWeight", "carbonPDW")

# Nitrogen & Phosphorus
N1 <- convertTotal2PW(trait.num, dw, "nitrogenPDW", "nitrogenTotal")
N2 <- convertPW2Total(trait.num, dw,  "nitrogenTotal", "nitrogenPDW")
P1 <- convertTotal2PW(trait.num, dw, "phosphorusPDW", "phosphorusTotal")
P2 <- convertPW2Total(trait.num, dw, "phosphorusTotal", "phosphorusPDW")

# Ash
ash1 <- convertTotal2PW(trait.num, dw, "ashPDW", "ashTotal") 
ash2 <- convertPW2Total(trait.num, dw, "ashTotal", "ashPDW") 

# dryweightPWW from wetWeight and dryWeight
dwpww <- convertTotal2PW(trait.num, ww, "dryWeightPWW","dryWeight") %>% 
  filter(taxonID %notin% filter(trait.num, traitName == "dryWeightPWW")$taxonID)
waterpww <- dwpww %>% 
  filter(taxonID %notin% filter(trait.num, traitName == "waterPWW")$taxonID) %>% 
  mutate(traitValue = 100 - traitValue, 
         basisOfRecord = "derived from related trait", isDerived = TRUE,
         catalogSource = catalogNumber, verbatimTraitName = traitName,
         verbatimTraitUnit = traitUnit, verbatimNotes = notes,
         traitName = "waterPWW", traitID = 19) %>% 
  mutate(observationNumber = NA, maxObsNum = NA)


derived.comp <- bind_rows(C1, C2, C3, N1, N2, P1, P2, 
                          ash1, ash2, waterpww, dwpww) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA)

trait.num <- trait.num %>% 
  bind_rows(derived.comp)
  # updateIDs()
```

## Elemental ratios
```{r}
rCN <- calculateRatio(trait.num, "ratioCN", "carbonWeight", "nitrogenTotal")
rCP <- calculateRatio(trait.num, "ratioCP", "carbonWeight", "phosphorusTotal")
rNP <- calculateRatio(trait.num, "ratioNP", "nitrogenTotal", "phosphorusTotal")

derived.ratio <- bind_rows(rCN, rCP, rNP) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA)

trait.num <- trait.num %>% 
  bind_rows(derived.ratio)
  # updateIDs()
```



## Weight-specific rates
These are for rate values with no associated size values or if associated size is different.
Note both N & P excretion rates dont require this.
```{r}
rate.R1.a <- convertRateInd2WS(trait.num, dw, "respirationRate_WSDW_15C",
                             "respirationRate_15C", "ul O2 mg DW^-1 h^-1")
rate.R1.b <- convertRateWS2Ind(trait.num, dw, "respirationRate_15C",
                               "respirationRate_WSDW_15C", "ul O2 h^-1")
rate.R3.a <- convertRateInd2WS(trait.num, cw, "respirationRate_WSC_15C","respirationRate_15C",
                     "ul O2 mg C^-1 h^-1")
rate.R3.b <- convertRateWS2Ind(trait.num, cw, "respirationRate_15C",  
                               "respirationRate_WSC_15C", "ul O2 h^-1") %>% 
  filter(taxonID %notin% rate.R1.b$taxonID) 
rate.R2.a <- convertRateInd2WS(trait.num, ww, "respirationRate_WSWW_15C",
                             "respirationRate_15C", "ul O2 mg WW^-1 h^-1")
rate.R2.b <- convertRateWS2Ind(trait.num, ww, "respirationRate_15C",  
                               "respirationRate_WSWW_15C", "ul O2 h^-1") %>% 
  filter(taxonID %notin% rate.R1.b$taxonID) %>% 
  filter(taxonID %notin% rate.R3.b$taxonID)

rate.R.b <- bind_rows(rate.R1.b, rate.R2.b, rate.R3.b) 


rate.C.a <- convertRateInd2WS(trait.num, cw, "clearanceRate_WSC_15C","clearanceRate_15C",
                     "ml mg C^-1 h^-1")
rate.C.b <- convertRateWS2Ind(trait.num, cw, "clearanceRate_15C", "clearanceRate_WSC_15C",
                     "ml h^-1")
rate.I.a <- convertRateInd2WS(trait.num, cw, "ingestionRate_WSC_15C","ingestionRate_15C",
                     "ug C mg C^-1 h^-1")
rate.I.b <- convertRateWS2Ind(trait.num, cw, "ingestionRate_15C","ingestionRate_WSC_15C",
                     "ug C h^-1")
rate.EN.a <- convertRateInd2WS(trait.num, dw, "excretionRateN_WSDW_15C","excretionRateN_15C",
                     "ug N-NH4+ mg DW^-1 h^-1")
rate.EN.b <- convertRateWS2Ind(trait.num, dw, "excretionRateN_15C","excretionRateN_WSDW_15C",
                     "ug N-NH4+ h^-1")
rate.EP.a <- convertRateInd2WS(trait.num, dw, "excretionRateP_WSDW_15C","excretionRateP_15C",
                     "ug P-PO43- mg DW^-1 h^-1")
rate.EP.b <- convertRateWS2Ind(trait.num, dw, "excretionRateP_15C","excretionRateP_WSDW_15C",
                     "ug P-PO43- h^-1")

# Most WS2Ind return empty dataframes but retain the code here so it is updatable.
derived.rates <- bind_rows(rate.R1.a, rate.R2.a, rate.R3.a, rate.R.b, 
                          rate.C.a, rate.I.a, rate.EN.a, rate.EP.a,
                          rate.C.b, rate.I.b, rate.EN.b, rate.EP.b) %>% 
  mutate(catalogNumber = NA, observationNumber = NA, maxObsNum = NA)

trait.num <- trait.num %>% 
  bind_rows(derived.rates) 
  # updateIDs()
```

# Finalize and filter traits
```{r}
trait.num <- trait.num %>% 
  # Remove rate traits not standardized to 15 deg C
  filter(traitName %notin% c("clearanceRate","clearanceRate_WSC",
                          "ingestionRate","ingestionRate_WSC",
                          "filtrationRate","filtrationRate_WSC",
                          "growthRate","growthRate_WSC",
                          "growthRatePLength",
                          "respirationRate","respirationRate_WSDW",
                          "respirationRate_WSWW","respirationRate_WSC",
                          "excretionRateN","excretionRateN_WSDW",
                          "excretionRateP","excretionRateP_WSDW")) 

traits.lvl2 <- trait.num %>% 
  mutate(traitValue = as.character(traitValue)) %>%
  bind_rows(trait.cat) %>% 
  # Remove cubozoans
  filter(majorgroup %notin% c("Cubozoan"))

# Exclude traits with N <= 30 taxa 
exclude.traits <- traits.lvl2 %>% 
  # filter(taxonRank %in% c("Species","Subspecies")) %>% 
  group_by(traitID, traitName, taxonID, scientificName, taxonRank) %>% 
  summarise(n = n()) %>% 
  group_by(traitID, traitName) %>% 
  summarise(n = n()) %>% 
  ungroup() %>% 
  filter(n <= 30)

traits.lvl2.excluded <- traits.lvl2 %>% 
  filter(traitID %in% exclude.traits$traitID) 

traits.lvl2 <- traits.lvl2 %>% 
  filter(traitID %notin% exclude.traits$traitID) %>% 
  # add binary traits
  bind_rows(trait.binary)
```


# Export to csv files
```{r, eval=FALSE}
# write.csv(traits.lvl2,
#           file = paste0(outfol,"trait_dataset_level2-",
#                    ymd(Sys.Date()),".csv"),
#           row.names = FALSE)
# write.csv(traits.lvl2.excluded,
#           file = paste0(outfol,"trait_dataset_level2_excluded-",
#                    ymd(Sys.Date()),".csv"), 
#           row.names = FALSE)
```



