---
title: "Finalizing level 2 dataset"
author: "Patrick Pata"
date: "01/12/2022"
output: html_document
---

This file provides the scripts to produce figures 3 and 5.

This file loads the level 2 trait dataset of species level averages with derived trait values and finalizes and exports the level 2 trait dataset. Rate traits that were not standardized to 15 degrees celsius, i.e., those that do not have an associated temperature value, are excluded. If there are other revisions to the level 2 trait dataset, these could be included here. The numerical, categorical, and binary traits are then merged into a single dataframe for the level 2 trait dataset.

The rest of the file includes codes that summarize the level 2 trait data in tables and figures. The data sources of the traits are catalogued and a summary table of traits and which references they were originally sourced from is created. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and functions
```{r}
library(tidyverse)
library(openxlsx)
library(lubridate)
library(ggbreak) # cite: 10.3389/fgene.2021.774846
library(cowplot)

source("toolkit.R")

# theme_set(theme_bw())
```

# Prepare files and folders
```{r}
s.format <- read.csv("data_input/trait_dataset_standard_format_20230628.csv")[-1,]

trait.directory <- read.csv("data_input/trait_directory_20230628.csv") %>% 
  distinct(traitID, .keep_all = TRUE)

# taxonomy table
taxonomy <- read.csv("data_input/taxonomy_table_20230628.csv") %>% 
  distinct(taxonID, .keep_all = TRUE)

# stage table
lifestagelist <- read.csv("data_input/lifestage_directory_20230628.csv") %>% 
  select(-c(majorgroup, notes))


infol <- "data_input/Trait_dataset_level2/"
outfol <- "data_input/Trait_dataset_level2/"

# Load Level 2 dataset
traits.lvl2 <- read.csv(paste0(infol,"trait_dataset_level2-2023-06-28.csv"))

# # Count number of data records
# traits.lvl2 %>% filter(valueType %in% c("numeric","categorical")) %>% 
#   distinct(traitID, taxonID) %>% nrow()
# 
# traits.lvl2 %>% filter(valueType %in% c("numeric","categorical")) %>% 
#   filter(taxonRank %in% c("Species","Subspecies")) %>% 
#   distinct(taxonID) %>% nrow()
```



# Figure 3. Summary of level 2 data coverage as the number of species records per trait by major group. 
Only traits recorded at the species level were enumerated.
```{r, fig.height=4, fig.width=3}

# Assign colors to major groups
majorgroup.colors <- data.frame(
  majorgroup = c("Calanoid","Non-calanoid","Mysid","Amphipod","Decapod","Euphausiid",
                 "Ostracod","Polychaete","Pteropod",
                 "Chaetognath","Appendicularian","Thaliacean","Cladoceran",
                 "Hydromedusae","Siphonophore","Scyphomedusae","Ctenophore"),
  color = c("blue","navy","slateblue2","cornflowerblue","cyan2","forestgreen",
            "seagreen3","tan1","green",
            "darkorchid4","violetred","hotpink3","saddlebrown",
            "tomato1","goldenrod3","yellow2","red4"))
            
# To slightly change the labels for the plot
majorgroup.colors.ed.labels <-  c("Calanoids","Non-calanoids","Mysids",
                                  "Amphipods","Decapods","Euphausiids",
                                  "Ostracods","Polychaetes","Pteropods",
                                  "Chaetognaths","Appendicularians","Thaliaceans",
                                  "Cladocerans","Hydromedusae",
                                  "Siphonophores","Scyphozoans","Ctenophores")

trait.sum <- traits.lvl2 %>% 
  filter(valueType %in% c("numeric","categorical")) %>% 
  dplyr::select(traitID, traitName, taxonID, scientificName, majorgroup, taxonRank,
         primaryReference, secondaryReference) %>% 
  left_join(distinct(trait.directory,
                     traitBucket, traitNum, trait, traitID),
            by = "traitID") %>% 
  # only consider species and subspecies level when presenting results
  filter(taxonRank %in% c("Species","Subspecies")) 

# Number of species with at least one trait info
length(unique(trait.sum$taxonID))
  
# count number of unique species-trait records (note that this is not traitName thus nitrogenPDW and nitrogenTotal are the same trait:nitrogen content)
trait.sum <- trait.sum %>% 
  distinct(traitBucket, traitNum, traitName, majorgroup, taxonID) %>% 
  group_by(traitBucket, traitNum, traitName, majorgroup) %>% 
  # Number of species-trait per major group
  summarise(Nsp.mg = n()) %>% 
  group_by(traitBucket, traitNum, traitName) %>% 
  # Number of species per trait
  mutate(Nspecies = sum(Nsp.mg)) %>% 
  arrange(traitNum) %>% 
  ungroup() %>% 
  # arrange by trait bucket and decreasing total value
  mutate(majorgroup = factor(majorgroup, levels = majorgroup.colors$majorgroup),
         traitBucket = factor(traitBucket, 
                              levels = c("morphological","biochemical composition",
                                         "physiological","behavioral",
                                         "life history"))) %>% 
  mutate(traitName = fct_reorder(traitName, Nspecies))

            
# Stacked barplot with count at end
bplot <- ggplot(trait.sum, aes(x = traitName, y = Nsp.mg, 
                               fill = majorgroup)) +
  
  geom_bar(stat = "identity") +
  # Set axis break
  scale_y_continuous(breaks = c(0, 250, 500, 750, 1000, 3000),
                     limits = c(0, 3180)) +
  scale_y_break(c(1250,2900), expand = FALSE) +
  # Add number at end of bar
  geom_text(aes(label = after_stat(y), group = traitName),
            stat = "summary", fun = sum, hjust = -0.1,
            size = 5) +
  scale_fill_manual(values = majorgroup.colors$color,
                    limits = majorgroup.colors$majorgroup,
                    labels = majorgroup.colors.ed.labels,
                    name = "Major group") +
  coord_flip() +
  xlab("Trait") + ylab("# Species") +
  # facets proportioned by number of traits
  facet_grid(traitBucket ~., scales = "free_y", space = "free", switch = "y")  +
  theme_bw() +
  theme(strip.placement = "outside", text = element_text(size = 20),
        legend.position = c(0.8, 0.5),
        legend.background = element_rect(fill = "white", colour = NA)) 
bplot 

ggsave(filename = "figures/fig_3_coverage_barplot_species.png",
    width = 32, height=37, units = "cm", dpi = 300)
```


## Figure 3 version enumerated by trait group
```{r, fig.height=3, fig.width=3, eval = FALSE}

# Assign colors to major groups
majorgroup.colors <- data.frame(
  majorgroup = c("Calanoid","Non-calanoid","Mysid","Amphipod","Decapod","Euphausiid",
                 "Ostracod","Polychaete","Pteropod",
                 "Chaetognath","Appendicularian","Thaliacean","Cladoceran",
                 "Hydromedusae","Siphonophore","Scyphomedusae","Ctenophore"),
  color = c("blue","navy","slateblue2","cornflowerblue","cyan2","forestgreen",
            "seagreen3","tan1","green2",
            "darkorchid4","violetred","hotpink3","saddlebrown",
            "tomato1","goldenrod3","yellow2","red4"))
            

trait.sum <- traits.lvl2 %>% 
  filter(valueType %in% c("numeric","categorical")) %>% 
  select(traitID, traitName, taxonID, scientificName, majorgroup, taxonRank,
         primaryReference, secondaryReference) %>% 
  left_join(distinct(trait.directory,
                     traitBucket, traitNum, trait, traitID),
            by = "traitID") %>% 
  # only consider species and subspecies level when presenting results
  filter(taxonRank %in% c("Species","Subspecies")) 

# Number of species with at least one trait info
length(unique(trait.sum$taxonID))
  
# count number of unique species-trait records (note that this is not traitName thus nitrogenPDW and nitrogenTotal are the same trait:nitrogen content)
trait.sum <- trait.sum %>% 
  distinct(traitBucket, traitNum, trait, majorgroup, taxonID) %>% 
  group_by(traitBucket, traitNum, trait, majorgroup) %>% 
  # Number of species-trait per major group
  summarise(Nsp.mg = n()) %>% 
  group_by(traitBucket, traitNum, trait) %>% 
  # Number of species per trait
  mutate(Nspecies = sum(Nsp.mg)) %>% 
  arrange(traitNum) %>% 
  ungroup() %>% 
  # arrange by trait bucket and decreasing total value
  mutate(majorgroup = factor(majorgroup, levels = majorgroup.colors$majorgroup),
         traitBucket = factor(traitBucket, 
                              levels = c("morphological","biochemical composition",
                                         "physiological","behavioral",
                                         "life history"))) %>% 
  mutate(trait = fct_reorder(trait, Nspecies))

            
# Stacked barplot with count at end
bplot <- ggplot(trait.sum, aes(x = trait, y = Nsp.mg, 
                               fill = majorgroup)) +
  geom_bar(stat = "identity") +
  # Set axis break
  scale_y_continuous(breaks = c(0, 250, 500, 750, 1000, 3000),
                     limits = c(0, 3180)) +
  scale_y_break(c(1250,2980), expand = FALSE) +
  # Add number at end of bar
  geom_text(aes(label = after_stat(y), group = trait),
            stat = "summary", fun = sum, hjust = -0.1,
            size = 3) +
  scale_fill_manual(values = majorgroup.colors$color,
                    limits = majorgroup.colors$majorgroup,
                    name = "Major group") +
  coord_flip() +
  xlab("Trait") + ylab("# Species") +
  # facets proportioned by number of traits
  facet_grid(traitBucket ~., scales = "free_y", space = "free", switch = "y") +
  theme_bw() +
  theme(strip.placement = "outside") 
bplot
```


# Figure 5. Distribution of values of eight traits for each of the zooplankton classes.
```{r, fig.width=3, fig.height=3, eval=TRUE, warning=FALSE}
scaleFUN.int <- function(x) sprintf("%.0f", x)

# Distribution of Size + nitrogen content and respiration rate
trait.num <- traits.lvl2 %>% 
  filter(valueType == "numeric") %>% 
  mutate(traitValue = as.numeric(traitValue))
# get mean size per major group to reorder majorgroups
A <- trait.num %>% 
  filter(majorgroup %notin% c("Cubozoan")) %>% 
  filter(traitName == "bodyLengthMax") %>% 
  group_by(class) %>% 
  summarise(meanlength = mean(traitValue), sd = sd(traitValue), 
            median = median(traitValue), n = n()) %>% 
  arrange(meanlength)

# Filter trait information and assign factors 
B <- trait.num %>% 
  filter(majorgroup %notin% c("Cubozoan")) %>% 
  filter(traitName %in% c("bodyLengthMax", "carbonWeight", "wetWeight",
                          "nitrogenPDW","respirationRate_WSC_15C")) %>% 
  mutate(traitName = factor(traitName)) %>% 
  mutate(traitName = factor(traitName, 
                            levels = c("bodyLengthMax", "wetWeight",
                                       "nitrogenPDW","respirationRate_WSC_15C"))) %>%
  filter(!is.na(class)) %>% 
  # add sample size
  group_by(traitName, class) %>% 
  mutate(Nspecies = n()) %>% 
  mutate(meanValue = mean(traitValue)) %>% 
  ungroup() %>% 
  # Assign blank column and relevel factors
  bind_rows(data.frame(traitName = "wetWeight", Nspecies = 0,
                       class = c("Branchiopoda","Appendicularia") )) %>% 
  bind_rows(data.frame(traitName = "respirationRate_WSC_15C", Nspecies = 0,
                       class = c("Branchiopoda","Appendicularia",
                                 "Gastropoda","Sagittoidea","Polychaeta") )) %>% 
  # Order by length
  mutate(class = factor(class, levels = A$class))
  


# By cowplot
j1 <- ggplot(filter(B, traitName == "bodyLengthMax"),
             aes(x = class, y = traitValue, fill = Nspecies)) +
  geom_violin() +  #scale="area", width=1.5
  geom_point(aes(y = meanValue), size =2, color = "black") +
  scale_y_continuous(trans = "log10", labels = scaleFUN.int) +
  scale_x_discrete(breaks = A$class) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right", 
        legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10)) +
  scale_fill_continuous("N species", type = "viridis", trans = "log10") +
  ylab("Maximum body length\n(mm)")


j2 <- ggplot(filter(B, traitName == "wetWeight"),
             aes(x = class, y = traitValue, fill = Nspecies)) +
  geom_violin() +  #scale="area", width=1.5
  geom_point(aes(y = meanValue), size =2, color = "black") +
  scale_y_continuous(trans = "log10") +
  scale_x_discrete(breaks = A$class) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right",
        legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10)) +
  scale_fill_continuous("N species", type = "viridis", trans = "log10") +
  ylab("Wet weight (mg)")


j3 <- ggplot(filter(B, traitName == "nitrogenPDW"),
             aes(x = class, y = traitValue, fill = Nspecies)) +
  geom_violin() +  #scale="area", width=1.5
  geom_point(aes(y = meanValue), size =2, color = "black") +
  scale_y_continuous(trans = "log10", labels = scaleFUN.int) +
  scale_x_discrete(breaks = A$class) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right",
        legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10)) +
  scale_fill_continuous("N species", type = "viridis", trans = "log10") +
  ylab("Nitrogen content \n(percent dry weight)")


j4 <- ggplot(filter(B, traitName == "respirationRate_WSC_15C"),
             aes(x = class, y = traitValue, fill = Nspecies)) +
  geom_violin() +
  geom_point(aes(y = meanValue), size =2, color = "black") +
  scale_y_continuous(trans = "log10", labels = scaleFUN.int) +
  scale_x_discrete(breaks = A$class) +
  theme_bw() +
  scale_fill_continuous("N species", type = "viridis", trans = "log10") +
  ylab("Respiration rate \n(ul O2 mg C^-1 h^-1)") + xlab("Class") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 12),
        legend.position = "right",
        legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10)) 


J <- plot_grid(j1, j2, j3, j4,
               ncol = 1, align = "v", rel_heights = c(1,1,1,1.6),
               labels = c("A","B","C","D"))

# Distribution of some common categorical traits
cat.traits <- traits.lvl2 %>% 
  filter(majorgroup %notin% c("Cubozoans"), 
         class %notin% c("Cubozoa")) %>% 
  # Order arbitrarily by broad phylogeny
  mutate(class = factor(class, levels = A$class))

# Visualize the number of species associated with a particular categorical trait
vertical.distribution <- cat.traits %>% 
  filter(traitName == "verticalDistribution") %>% 
  group_by(traitValue, class) %>% 
  summarise(Nspecies = n()) %>% 
  rename(traitName = traitValue) %>% 
  mutate(traitName = factor(traitName,
                             levels = c("epipelagic","epimesopelagic",
                                        "epibathypelagic","mesopelagic",
                                        "mesobathypelagic","bathypelagic"))) %>% 
  ungroup() %>% 
  mutate(trait = "Vertical Distribution")

repro.mode <- cat.traits %>% 
  filter(traitName %in% c("RM.broadcasting","RM.brooding","RM.asexual"),
         traitValue == 1) %>% 
  # Categorize asexual into brooding group
  mutate(traitName = if_else(traitName == "RM.asexual",
                             "RM.brooding",traitName)) %>% 
  # rename trait levels by removing "RM." prefix
  mutate(traitName = str_replace(traitName, "RM.","")) %>% 
  group_by(traitName, class) %>% 
  summarise(Nspecies = n()) %>% 
  ungroup() %>% 
  # Assign blank column and relevel factors
  bind_rows(data.frame(traitName = "brooding", class = "Scyphozoa",
                       Nspecies = 0)) %>% 
  mutate(class = factor(class, levels = A$class)) %>% 
  mutate(trait = "Reproduction Mode")

feeding.mode <- cat.traits %>% 
  filter(traitName %in% c("FM.passive.ambush","FM.active.ambush","FM.cruise",
                          "FM.current","FM.particle.feeder"),
         traitValue == 1) %>% 
  mutate(traitName = str_replace(traitName, "FM.","")) %>% 
  group_by(traitName, class) %>% 
  summarise(Nspecies = n()) %>% 
  ungroup() %>% 
  # Assign blank column and relevel factors
  # bind_rows(data.frame(traitName = "cruise", majorgroup = "Decapod",
  #                      Nspecies = 0)) %>% 
  mutate(class = factor(class, levels = A$class),
         traitName = factor(traitName, 
                            levels = c("cruise","current","active.ambush",
                                       "passive.ambush","particle.feeder"))) %>% 
  mutate(trait = "Feeding Mode")
  
trophic.group <- cat.traits %>% 
  filter(traitName %in% c("TG.omnivore","TG.herbivore",
                          "TG.carnivore","TG.detritivore"),
         traitValue == 1) %>% 
  mutate(traitName = str_replace(traitName, "TG.","")) %>% 
  group_by(traitName, class) %>% 
  summarise(Nspecies = n()) %>% 
  ungroup() %>% 
  mutate(traitName = factor(traitName, 
                            levels = c("herbivore","omnivore",
                                       "carnivore","detritivore"))) %>% 
  mutate(trait = "Trophic Group") 
  

h1 <- ggplot(vertical.distribution,
             aes(x = class, y = Nspecies, fill = traitName)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent, position = "left", expand = c(0, 0))  + 
  xlab("Major group") + ylab("% Records") +
  scale_x_discrete(breaks = A$class, expand = c(0, 0)) +
  scale_fill_manual("Vertical Distribution",
                    values = c("turquoise3", "seagreen4", "darkgreen",
                              "mediumblue","purple4","gray10")) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10))

h2 <- ggplot(repro.mode,
             aes(x = class, y = Nspecies, fill = traitName)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent, position = "left", expand = c(0, 0))  + 
  xlab("Major group") + ylab("% Records") +
  scale_x_discrete(breaks = A$class, expand = c(0, 0)) +
  scale_fill_manual("Reproduction Mode",
    values = c("lightpink","lightskyblue")) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10))

h3 <- ggplot(feeding.mode,
             aes(x = class, y = Nspecies, fill = traitName)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent, position = "left", expand = c(0, 0))  + 
  xlab("Major group") + ylab("% Records") +
  scale_x_discrete(breaks = A$class, expand = c(0, 0)) +
  scale_fill_manual("Feeding Mode",
    values = c("maroon4","hotpink3","khaki3","tan4","steelblue4")) +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "right", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10))

h4 <- ggplot(trophic.group,
             aes(x = class, y = Nspecies, fill = traitName)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent, position = "left", expand = c(0, 0))  + 
  xlab("Class") + ylab("% Records") +
  scale_fill_manual("Trophic Group",
    values = c("olivedrab3","darkorange","firebrick3","darkorchid4")) +
  scale_x_discrete(breaks = A$class, expand = c(0, 0)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 12),
        legend.position = "right", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(0,0,0,-10)) 

H <- plot_grid(h1, h2, h3, h4, 
               ncol = 1, align = "v", rel_heights = c(1,1,1,1.6),
               labels = c("E","F","G","H"))


# Merge summaries into a single figure
fig <- plot_grid(J, H, nrow = 1, rel_widths = c(1,1.1))

fig

ggsave(filename = "figures/fig_5_trait_distributions.png",
    width = 24.75, height=30, units = "cm", dpi = 300)
```



# Supp Figure 2. Size bias in trait coverage. 
```{r}
# Data frame of body lengths
A.length <- traits.lvl2 %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(valueType %in% c("numeric","categorical")) %>% 
  filter(traitName == "bodyLengthMax") %>% 
  mutate(traitValue = as.numeric(traitValue)) %>% 
  select(taxonID, scientificName, bodyLength = traitValue)

# Data frame counting number of trait records categorize by "trait" and not trait name
B.all.traits <- traits.lvl2 %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(valueType %in% c("numeric","categorical"))  %>% 
  filter(traitName != "bodyLengthMax") %>% 
  # Add "trait" info by joining with trait directory
  left_join(select(trait.directory, traitName, trait)) %>%   
  distinct(taxonID, scientificName, valueType, trait,  phylum, class) %>% 
  group_by(taxonID, scientificName, valueType, phylum, class) %>% 
  summarise(Ntraits = n()) %>% 
  pivot_wider(names_from = valueType, values_from = Ntraits, values_fill = 0)

C2 <- inner_join(B.all.traits, A.length, by = c("taxonID","scientificName"))

ggplot(filter(C2, numeric > 0), 
       aes(x = bodyLength, y = numeric)) +
  geom_point(aes(color = phylum), alpha = 0.3) +
  geom_smooth(aes(color = phylum), method = "lm", se = FALSE) +
  scale_color_manual(values = c("tan1","blue","darkorchid4","violetred",
                                      "goldenrod3","red4","green2"),
  name = "Phylum") +
  geom_smooth(data = C2, aes(x = bodyLength, y = numeric), 
              color = "black") +
  xlab("Body length (mm)") + ylab("Number of numerical traits") +
  scale_x_continuous(trans = "log10")

ggsave("figures/supp_fig_2_bias_numtraits_size.png",
       width = 7.5, height = 3, units = "in", dpi = 300)
```
