---
title: "Taxonomic coverage at level 2 data"
author: "Patrick Pata"
date: '2023-01-11'
output: html_document
---

This file provides the scripts to produce supplementary figure 1 and the input files for making figure 4. Summary of level 2 data coverage as the number of trait records per species. The phylogenetic dendrogram and trait distribution figure was generated in the iTOL website and annotated in photoshop. This script thus demonstrates how the input files to generate the figure are created.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and functions
```{r}
library(tidyverse)
library(gplots) # for converting color names to hex format
library(phyr)
library(rotl) # for phylo distance

source("toolkit.R")
```

# Prepare files and folders
```{r}
s.format <- read.csv("data_input/trait_dataset_standard_format_20230628.csv")[-1,]

trait.directory <- read.csv("data_input/trait_directory_20230628.csv") %>% 
  distinct(traitID, .keep_all = TRUE)

# taxonomy table
taxonomy <- read.csv("data_input/taxonomy_table_20230628.csv") %>% 
  distinct(taxonID, .keep_all = TRUE)

# stage table
lifestagelist <- read.csv("data_input/lifestage_directory_20230628.csv") %>% 
  select(-c(majorgroup, notes))


infol <- "data_input/Trait_dataset_level2/"
outfol <- "data_input/Trait_dataset_level2/"

# Load Level 2 dataset
traits.lvl2 <- read.csv(paste0(infol,"trait_dataset_level2-2023-06-28.csv"))

# The zooplankton phylogenetic tree file extracted from the Tree of Life database and subset for species found in the zooplankton trait dataset.
# load("data_input/phylo_tree_trait_level2_20230501.RData")
```

# Supp Figure 1. Distribution of the total number of traits per species.
Use the curated level 2 data as basis for taxonomic converage
```{r, warning=FALSE}
# Create dataframe of scientific names at species level for data at trait level 2 (will have same list as level 1) to list number of traits per species. This dataframe will be used in searching for ott_ids.
taxon.count <- traits.lvl2 %>% 
  filter(valueType %notin% c("binary")) %>% 
  group_by(taxonID) %>% 
  count() %>% 
  ungroup() %>% 
  arrange(-n) %>% 
  left_join(distinct(taxonomy, taxonID, .keep_all = T ), by = "taxonID") %>% 
  # only plot subspecies and species level
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(majorgroup != "Cubozoan") %>% 
  filter(!is.na(aphiaID)) %>% 
  # for subpecies and names with parentheses, remove these
  mutate(species = gsub(r"{\s*\([^\)]+\)}","",as.character(scientificName))) %>% 
  # remove duplicated subspecies name for ott search
  # mutate(species = paste(unique(unlist(strsplit(species, split = " "))),
  #                        collapse = " ")) %>%
  mutate(species = word(species,1,2)) %>%
  relocate(taxonID,species) %>% 
  # Prepare for ott search
  mutate(search_string = tolower(species)) %>% 
  dplyr::select(-c(verbatimScientificName))

# Plot distribution of number of trait information per taxa
taxon.count.sum <- taxon.count %>% 
  rename(ntraits = n) %>% 
  group_by(ntraits) %>% 
  summarise(ntaxa = n()) %>% 
  ungroup() %>% 
  mutate(cum = cumsum(ntaxa) ) %>% 
  mutate(perc = cum/ sum(ntaxa) * 100) %>% 
  mutate(perc.rev = 100 - perc)

# For y axis scaling
ylim.prim <- c(0, 2000)   # in this example, precipitation
ylim.sec <- c(0, 100)
b <- diff(ylim.prim)/diff(ylim.sec)
a <- ylim.prim[1] - b*ylim.sec[1]

ggplot(taxon.count.sum, aes(ntraits, ntaxa)) +
  geom_point() +
  theme_bw() +
  geom_segment( aes(x=ntraits, xend=ntraits, y=0, yend=ntaxa)) +
  geom_line(aes(y = a + perc*b), linetype = "dashed") +
  scale_y_continuous("Number of species", 
                     sec.axis = sec_axis(~ (. -a)/b, name = "% of species")) + 
  scale_x_continuous("Number of traits per species",
                     breaks = c(0,10,20,30,40,50,60))

ggsave(filename = "figures/supp_fig_1_num_traits_by_species.png",
    width = 7, height=3, units = "in", dpi = 300)
```

# For Figure 4. Summary of level 2 data coverage as the number of trait records per species.
## Convert taxonomy table to phylogenetic tree
```{r, warning=FALSE}
resolved_names <- rotl::tnrs_match_names(taxon.count$species, 
                                   context_name = "Animals") %>% 
  # ott ids removed because does not match the actual species, 
  filter(ott_id %notin% c(3681135,2875111,1026631,3589733)) %>% 
  # exclude ott ids that prevent tree builing
  filter(ott_id %notin% c(2946383,2946644,2951566,2957044,3589733,3681127,3681811,
            3683799,4669348,4712943,4984796,5092841,5296504,5520904,615879,3683796,
            624879,7489044,7839149,787131))
  

taxon.count <- taxon.count %>% 
  left_join(dplyr::select(resolved_names, search_string, ott_id),
            by = "search_string") %>% 
  # Mesopodopsis sp is in wrong part of tree so exclude for now
  filter(genus %notin% c("Mesopodopsis")) %>% 
  filter(!is.na(ott_id)) %>% 
  relocate(ott_id, species, n, majorgroup) %>% 
  distinct(ott_id, .keep_all = TRUE)
  

# order by major group
taxon.count <- taxon.count %>% 
  arrange(phylum,class,order,family,genus)

# Create the zooplankton phylogenetic tree based on the curated taxon list
zoop_tree <- rotl::tol_induced_subtree(ott_ids = taxon.count$ott_id)

# Barneche et al. (2018): Derive variance-covariance matrix based on Brownian evolution from the phylogenetic tree. If branch length are unknown, set to length equal to number of descendant tips minus one (arbitrary method of Grafen 1989).
# compute branch lengths
zoop_tree <- ape::compute.brlen(zoop_tree, method = 'Grafen')


# Assign colors to major groups
majorgroup.colors <- data.frame(
  majorgroup = c("Calanoid","Non-calanoid","Mysid","Amphipod","Decapod",
                 "Euphausiid","Ostracod","Polychaete","Pteropod",
                 "Chaetognath","Appendicularian","Thaliacean","Cladoceran",
                 "Hydromedusae","Siphonophore","Scyphomedusae","Ctenophore"),
  color = c("blue","navy","slateblue2","cornflowerblue","cyan2","forestgreen",
            "seagreen3","tan1","green2",
            "darkorchid4","violetred","hotpink3","saddlebrown",
            "tomato1","goldenrod3","yellow2","red4"))


# Create table of tips with updated labels and grouping for color
tree.tips <- data.frame(label = zoop_tree$tip.label) %>% 
  mutate(ott_id = as.numeric(sub(".*ott","",label))) %>% 
  left_join(dplyr::select(taxon.count, ott_id, species, n, majorgroup,
                          phylum, class), by = "ott_id") %>% 
  left_join(majorgroup.colors, by = "majorgroup")

# # Save the phylo tree object
# save(zoop_tree, tree.tips, taxon.count,
#      file = "data_output/phylo_tree_trait_level2.RData")
```


## Prepare files for annotation in iTOL
Create files which will be loaded into iTOL (https://itol.embl.de/) where the figure will be generated. The figure is then annotated in photoshop to add zooplankton icons and percent trait data coverage.
```{r}
zoop_tree_export <- zoop_tree
zoop_tree_export$tip.label <-  sub(".*ott","", zoop_tree$tip.label)
zoop_tree_export$node.label <- sub(".*ott","", zoop_tree$node.label)

# Write tree to file for iTOL
ape::write.tree(zoop_tree_export, file='data_output/zoop_tree_species_1trait.txt')

# Export to labels to LABELS.txt template file
writeLines(c("LABELS","SEPARATOR COMMA","","DATA"), 
           "data_output/labels_template.txt")
# labels_template.txt
# NODE_ID, LABEL, CLASS
labels <- tree.tips %>% 
  dplyr::select(ott_id, species, majorgroup)

write.table(labels, file="data_output/labels_template.txt",
            append = TRUE, sep = ",",
            quote = FALSE, row.names = FALSE, col.names = FALSE)

# Export label style to DATASET_STYLE.txt template file
writeLines(c("DATASET_STYLE","SEPARATOR COMMA","",
             "DATASET_LABEL,Color labels by majorgroup","",
             "COLOR,#ffff00","",
             "DATA"), 
           "data_output/dataset_style_template.txt")

# ID,TYPE,WHAT,COLOR,WIDTH_OR_SIZE_FACTOR,STYLE,BACKGROUND_COLOR
styletab <- tree.tips %>% 
  dplyr::select(ott_id, color) %>% 
  mutate(color = gplots::col2hex(color)) %>% 
  mutate(type = "label", what = "node", textcolor = "#000000",
         fontsize = "3", style = "italic",
         #can specific background color
         bgcolor = gplots::col2hex(color)) %>%  
  dplyr::select(-color)

write.table(styletab, file = "data_output/dataset_style_template.txt",
            append = TRUE, sep = ",",
            quote = FALSE, row.names = FALSE, col.names = FALSE)


# COLOR STRIPS by major group
writeLines(c("DATASET_COLORSTRIP","SEPARATOR COMMA","",
             "DATASET_LABEL,Color strip by majorgroup","",
             "COLOR,#7FFFD4","",
             "COLOR_BRANCHES 1",
             "COMPLETE_BORDER 1",
             "SHOW_STRIP_LABELS 1",
             "STRIP_WIDTH 40",
             "SHOW_LABELS 0",
             "DATA"), 
           "data_output/dataset_colorstrip_template.txt")
striptab <- tree.tips %>% 
  dplyr::select(ott_id,color,majorgroup) %>% 
  mutate(color = gplots::col2hex(color))


write.table(striptab, file = "data_output/dataset_colorstrip_template.txt",
            append = TRUE, sep = ",",
            quote = FALSE, row.names = FALSE, col.names = FALSE)


# Annotate with a simple bar chart of N trait records
writeLines(c("DATASET_SIMPLEBAR","SEPARATOR COMMA","",
             "DATASET_LABEL,Number of trait records","",
             "COLOR,#666666","",
             "DATA"), 
           "data_output/dataset_simplebar_template.txt")

bartab <- tree.tips %>% 
  dplyr::select(ott_id,n,color) %>% 
  mutate(color = gplots::col2hex(color)) 

write.table(bartab, file = "data_output/dataset_simplebar_template.txt",
            append = TRUE, sep = ",",
            quote = FALSE, row.names = FALSE, col.names = FALSE)
```


## Calculate the percent trait coverage per major group using in annotating Figure 4
```{r}
major.group.summary <- taxon.count %>% 
  group_by(majorgroup) %>% 
  mutate(Nspecies = n()) %>% 
  # % more than 1 trait
  mutate(perc.1 = if_else(n > 1, 1, 0),
         perc.1 = sum(perc.1)/Nspecies*100,
         # % more than 5 traits
         perc.5 = if_else(n > 5, 1, 0),
         perc.5 = sum(perc.5)/Nspecies*100,
         # % more than 10 traits
         perc.10 = if_else(n > 10, 1, 0),
         perc.10 = sum(perc.10)/Nspecies*100,
         nmax = max(n)) %>% 
    distinct(majorgroup, Nspecies, perc.1, perc.5, perc.10, nmax) %>% 
  mutate(perc.5.round = round(perc.5)) %>% 
  relocate(majorgroup, Nspecies, perc.5.round, nmax)
```


# Open a NEXUS file exported from iTOL
```{r, eval = FALSE, fig.width = 5, fig.height = 5}
# nexus <- ape::read.nexus("data_output/phylo_tree/itol_tree_nexus.txt")
# ape::plot.phylo(nexus,  show.tip.label = FALSE)
```



