---
title: "phyR and Cramer's V analysis"
author: "Patrick Pata"
date: '2023-01-17'
output: html_document
---


This file provides the scripts to calculte the phylogenetic correlation between numerical traits and the Cramer's V analysis between categorical traits. Figure 6. Association between traits. is then generated.

The script for calculing the phylogenetic correlations are given but are not evaluated by default since this may take a bit of time. To generate the figure, the previously calculated correlations are loaded.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and functions
```{r}
library(tidyverse)
library(lubridate)
library(phyr)
library(rcompanion) # For Craymer's V
library(cowplot)

source("toolkit.R")
source("toolkit_analysis.R")
```


# Prepare files and folders
```{r}
s.format <- read.csv("data_input/trait_dataset_standard_format_20230628.csv")[-1,]

trait.directory <- read.csv("data_input/trait_directory_20230628.csv") %>% 
  distinct(traitID, .keep_all = TRUE)

# taxonomy table
taxonomy <- read.csv("data_input/taxonomy_table_20230628.csv") %>% 
  distinct(taxonID, .keep_all = TRUE)

# stage table
lifestagelist <- read.csv("data_input/lifestage_directory_20230628.csv") %>% 
  select(-c(majorgroup, notes))


infol <- "data_input/Trait_dataset_level2/"
outfol <- "data_input/Trait_dataset_level2/"

# Load Level 2 dataset
traits.lvl2 <- read.csv(paste0(infol,"trait_dataset_level2-2023-06-28.csv"))

# Consider all species-level traits and get the phylogenetic tree for these species.
load("data_input/phylo_tree_trait_level2_20230501.RData")
```

# Numerical trait correlation with phylogenetic effects
## Prepare the numeric trait dataset
Subset the database to only include values at the species level and apply necessary data transformations.
```{r, fig.height=10, fig.width=10}
trait.num <- traits.lvl2 %>% 
  filter(valueType == "numeric") %>% 
  # only analyze species/subspecies-level
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  mutate(traitValue = as.numeric(traitValue))

# Select which traits for correlation
trait.num.summary <- trait.num %>% 
  group_by(traitName) %>% 
  summarise(Nspecies = n()) %>% 
  # remove traits with <= 30 species
  filter(Nspecies > 30) %>% 
  arrange(-Nspecies)

# Filter the subset data to only traits which we have a good number of species to do correlations.
trait.num <- trait.num %>% 
  filter(traitName %in% trait.num.summary$traitName)

sort(unique(trait.num$traitName))


# Check data distributions and identify which need transformation

# Only energyDensityVolume does not need transformation. Log10 transformation results in normal distributions for 14 of the 45 traits, but still good to keep log10 transformation for the other traits anyway. 
# Percent composition traits would be better transformed with logit for regressions.
# Measures of size, rates, and ratios are okay for log10 transformation
trans.list <- trait.num %>% 
  distinct(traitName) %>% 
  mutate(transformation = "log10") %>% 
  mutate(transformation = if_else(traitName %in% c("energyDensityVolume",
                                                   "energyDensityWW",
                                                   "ingestionRate_WSC_15C"),
                                  "none",transformation),
         transformation = if_else(grepl("PDW|PWW|PAFDW",traitName),
                                  "logit",transformation))

# Loop through all variables and test for normality, additionally calculate the same if log transformed. (p-value < 0.05 is not normal distribution). Kolmogorov-Smirnov test is supposedly more reliable for high sample sizes but it does not handle ties in the data.
normal.test <- data.frame(traitName = character(), transformation = character(),
                          N = numeric(), SW = numeric(), SW.trans = numeric())
for(i in 1:nrow(trans.list)) {
  A <- filter(trait.num, traitName == trans.list[i,]$traitName)$traitValue
  normal.test[i,]$traitName <- trans.list[i,]$traitName
  normal.test[i,]$transformation = trans.list[i,]$transformation
  normal.test[i,]$N <- length(A)
  # Shapiro-Wilk test (at high sample size ~>50 this test becomes unreliable)
  normal.test[i,]$SW <- as.numeric(shapiro.test(A)$p.value)
  if (trans.list[i,]$transformation == "logit") { 
    normal.test[i,]$SW.trans <- as.numeric(shapiro.test(logitTransform(A))$p.value)
  } else {
    normal.test[i,]$SW.trans <- as.numeric(shapiro.test(log10(A))$p.value)
  }
}

normal.test <- normal.test %>% 
  mutate(isNorm.SW = if_else(SW >= 0.05, TRUE, FALSE),
         isNorm.SW.log = if_else(SW.trans >= 0.05, TRUE, FALSE),
         improvedSW = if_else(SW.trans > SW, TRUE, FALSE)) %>% 
  arrange(improvedSW)

for.trans <- normal.test %>% 
  filter(isNorm.SW == FALSE & improvedSW == TRUE) %>% 
  arrange(-SW.trans)
not.trans <- normal.test %>% 
  filter(isNorm.SW == TRUE | improvedSW == FALSE) %>% 
  arrange(-SW.trans)

# Finalize transformations for each trait
trans.list <- trans.list %>% 
  mutate(transformation = if_else(traitName %in% not.trans$traitName,
                                  "none", transformation))

# # Plots to visualize data transformation
# # Plot histogram distributions
# ggplot(data = trait.num, aes(x = traitValue)) +
#   geom_histogram() + 
#   facet_wrap(~traitName, nrow = 9, scales = "free")
# 
# # Q-Q plot of all untransformed data
# ggplot(data = trait.num, aes(sample = traitValue)) +
#    stat_qq() + stat_qq_line() +
#   facet_wrap(~traitName, nrow = 9, scales = "free")
#   
# # qqplot for traits to not transform
# ggplot(data = filter(trait.num, traitName %in% not.trans$traitName), 
#        aes(sample = traitValue)) +
#    stat_qq() + stat_qq_line() +
#   facet_wrap(~traitName, nrow = 2, scales = "free") +
#   ggtitle("Untransformed data")
#   
# # qqplot for transformed and logit
# ggplot(data = filter(trait.num, traitName %in% 
#                        filter(for.trans, transformation == "logit")$traitName), 
#        aes(sample = logitTransform(traitValue))) +
#    stat_qq() + stat_qq_line() +
#   facet_wrap(~traitName, nrow = 2, scales = "free") +
#   ggtitle("logit transformed data")
# 
# # qqplot for transformed and log10
# ggplot(data = filter(trait.num, traitName %in% 
#                        filter(for.trans, transformation == "log10")$traitName), 
#        aes(sample = log10(traitValue))) +
#    stat_qq() + stat_qq_line() +
#   facet_wrap(~traitName, nrow = 5, scales = "free") +
#   ggtitle("log10 transformed data")



# Prepare the phylogenetic data frame from correlation. Make sure to load the phylo tree from the previous script file, and confirm if it has sufficient coverage for this analysis.

# This will remove the "ott" in the tree tip label.
zoop_tree$tip.label <-  sub(".*ott","", zoop_tree$tip.label)

# Rename taxon.count and reorder based on tree.tips/zoop_tree labels
phylo.table <- tree.tips %>% 
  dplyr::select(ott_id, color) %>% 
  left_join(taxon.count, by = "ott_id")
rm(taxon.count, tree.tips)

# Subset the trait data frame with traits for analysis.
trait.num.phylo <- trait.num %>% 
  filter(taxonID %in% phylo.table$taxonID) %>% 
  
# Apply transformations to the data
  left_join(trans.list, by = "traitName")
# Log 10 transformation
ii <- which(trait.num.phylo$transformation == "log10")
trait.num.phylo[ii,]$traitValue = log10(trait.num.phylo[ii,]$traitValue)
# Logit transformation
ii <- which(trait.num.phylo$transformation == "logit")
trait.num.phylo[ii,]$traitValue = logitTransform(trait.num.phylo[ii,]$traitValue)

```


## Calculate correlations with phylogenetic effect
```{r, warning = FALSE, eval=FALSE}
# Identify and order which traits to correlate
trait.size <- c("bodyLengthMax","carbonWeight","dryWeight","wetWeight")
trait.water <- c("waterPWW","dryWeightPWW","carbonPWW")
trait.comp <- c("carbonPDW","ashTotal","ashPDW","afdwPDW",
                "nitrogenTotal","nitrogenPDW", "phosphorusTotal","phosphorusPDW",
                "carbohydratePDW","lipidPDW","lipidPWW",
                "proteinPDW","proteinPWW",
                "ratioCN","ratioCP","ratioNP","ratioProteinLipid",
                "energyContent","energyDensityDW",
                "energyDensityWW","energyDensityVolume")
trait.rate <- c("respirationRate_15C", "respirationRate_WSC_15C",
                "respirationRate_WSDW_15C","respirationRate_WSWW_15C",
                "excretionRateN_15C","excretionRateN_WSDW_15C",
                "excretionRateP_15C","excretionRateP_WSDW_15C",
                "clearanceRate_15C","clearanceRate_WSC_15C",
                "ingestionRate_15C","ingestionRate_WSC_15C",
                "growthRate_15C","growthRate_WSC_15C")
trait.repro <- c("developmentDuration",
                 "eggDiameter","eggWeight","clutchSize","fecundity")
trait.list <- c(trait.size, trait.water, trait.comp, trait.rate, trait.repro)

# Check if any traits are missing
traits.lvl2 %>% filter(valueType == "numeric") %>%  distinct(traitName) %>% 
  filter(traitName %notin% trait.list)

# Prepare a list of pairs of traits to correlate
trait.cor <- data.frame(trait1 = character(), trait2 = character())
for (i in c(1:(length(trait.list)-1))) {
  for (j in c(i:length(trait.list))) {
    trait.cor <- trait.cor %>% 
      bind_rows(data.frame(trait1 = trait.list[i],
                           trait2 = trait.list[j]))
  }
}
trait.cor <- trait.cor %>% 
  filter(trait1 != trait2)

# Loop through correlations and add test.num (~600 tests ~30 mins)
cor.phylo.tests <- data.frame()
for (i in c(1:nrow(trait.cor))){ #
  results <- corphylo.tests(trait.num.phylo, 
                            c(trait.cor$trait1[i], trait.cor$trait2[i]), 
                            zoop_tree,
                            logtransX = FALSE, logtransY = FALSE)
  if (is.data.frame(results)) {
    results <- results %>%  mutate(test.num = i)
    
    cor.phylo.tests <- cor.phylo.tests %>% 
      bind_rows(results)
  }
}

# # Export results
# save(cor.phylo.tests, file = "data_output/corphylo_20230410_traitmatrix.RData")
```

## Organize results of correlation tests
On average, the with phylo effects model results in a stronger correlation and ranks better as a model compared to a no phylo effects model (both with no ME).
```{r}
# Load the results of the chunk above
load("data_output/corphylo_20230410_traitmatrix.RData")

# Prepare a summary table from the cor phylo test results
B1 <- cor.phylo.tests %>% 
  distinct(analysis, model.type, cor.phylo) %>% 
  pivot_wider(names_from = model.type, values_from = cor.phylo) %>% 
  rename(r.star = "no.phylo.no.me", r.phylo = "phylo.no.me") %>% 
  # calculate change r
  mutate(r.change = r.phylo - r.star)
  # mutate(r.abs.change = abs(r.phylo) - abs(r.star)) # not a good measure for r~0
B2 <- cor.phylo.tests %>% 
  distinct(analysis, model.type, AIC) %>% 
  pivot_wider(names_from = model.type, values_from = AIC) %>% 
  rename(AIC.star = "no.phylo.no.me", AIC.phylo = "phylo.no.me") %>% 
  # calculate change in AIC
  mutate(AIC.change = AIC.phylo - AIC.star) %>%  # more negative, better
  mutate(is.phylo.better = if_else(AIC.phylo < AIC.star, TRUE, FALSE))
B <- cor.phylo.tests %>% 
  distinct(test.num, analysis, trait1, trait2, N.pairs, perc.cal) %>% 
  left_join(B1, by = "analysis") %>% 
  left_join(B2, by = "analysis") %>% 
  relocate(test.num)

C <- cor.phylo.tests %>% 
  filter(model.type == "phylo.no.me") %>% 
  mutate(trait.coef = rep(c("trait1","trait2"), times = cc))
C1 <- C %>% 
  distinct(analysis, trait.coef, d) %>% 
  mutate(d = as.numeric(d)) %>% 
  pivot_wider(names_from = trait.coef, values_from = d) %>% 
  rename(d.trait1 = trait1, d.trait2 = trait2)
C2 <- C %>% 
  distinct(analysis, trait.coef, B) %>% 
  mutate(B = as.numeric(B)) %>% 
  pivot_wider(names_from = trait.coef, values_from = B) %>% 
  rename(B.trait1 = trait1, B.trait2 = trait2)
C3 <- C %>% 
  distinct(analysis, trait.coef, pval) %>% 
  mutate(pval = as.numeric(pval)) %>% 
  pivot_wider(names_from = trait.coef, values_from = pval) %>% 
  rename(pval.trait1 = trait1, pval.trait2 = trait2)

cor.table <- B %>% 
  left_join(C1, by = "analysis") %>% 
  left_join(C2, by = "analysis") %>% 
  left_join(C3, by = "analysis") %>% 
  left_join(distinct(cor.phylo.tests, analysis, list), by = "analysis")
rm(B1, B2, B, cc, C, C1, C2, C3)

# Exclude correlations with <10 N.pairs (~5% of max N.pairs)
cor.table <- cor.table %>% 
  filter(N.pairs > 10) 
```

##  Plot correlation matrix figure 
The colors represent the correlation coefficient, the size represents sample size, and points are marked by a black circle to indicate statistical significance.
```{r, fig.height=5, fig.width=5}
cor.table.2 <- cor.table %>% 
  bind_rows(dplyr::select(cor.table, trait2 = trait1, trait1 = trait2, everything())) %>% 
   mutate(trait1 = factor(trait1, levels = trait.list),
         trait2 = factor(trait2, levels = trait.list)) %>% 
  mutate(trait1.num = as.numeric(trait1),
         trait2.num = as.numeric(trait2)) %>% 
  # retain only one side of the matrix
  arrange(trait1, trait2) %>% 
  group_by(analysis) %>% 
  filter(row_number() == 1) %>% 
  ungroup() %>% 
# How many correlation coefficients >0.50 and N.pairs > 100
  mutate(r.mag = if_else(abs(r.phylo) > 0.5 & N.pairs >= 0,
         "s","w")) %>% 
  mutate(N.species = N.pairs)
  
# Boxes to highlight parts of the correlation matrix
green.box <- data.frame(x=c(0.5,4.5,4.5,0.5), y=c(0.5,0.5,47.5,47.5))
yellow.box <- data.frame(x = c(4.5,7.5,7.5,4.5,   8.5,11.5,11.5,8.5, 
                               11.5,13.5,13.5,11.5, 13.5,15.5,15.5,13.5,
                               15.5,20.5,20.5,15.5,  20.5,23.5,23.5,20.5,
                               23.5,28.5,28.5,23.5,  28.5,32.5,32.5,28.5,
                               32.5,36.5,36.5,32.5,  36.5,40.5,40.5,36.5,
                               40.5,43.5,43.5,40.5,  43.5,47.5,47.5,43.5),
                         y = c(4.5,4.5,7.5,7.5,   8.5,8.5,11.5,11.5,
                               11.5,11.5,13.5,13.5,  13.5,13.5,15.5,15.5,
                               15.5,15.5,20.5,20.5,  20.5,20.5,23.5,23.5,
                               23.5,23.5,28.5,28.5,  28.5,28.5,32.5,32.5,
                               32.5,32.5,36.5,36.5,  36.5,36.5,40.5,40.5,
                               40.5,40.5,43.5,43.5,  43.5,43.5,47.5,47.5),
                         label = c(1,1,1,1, 2,2,2,2, 3,3,3,3, 4,4,4,4,
                                   5,5,5,5,  6,6,6,6, 7,7,7,7,  8,8,8,8,
                                   9,9,9,9,  10,10,10,10,  11,11,11,11,
                                   12,12,12,12))
box.labels <- data.frame(x = c(0.5,4.5,8.5,11.5,13.5,15.5,20.5,23.5,28.5,32.5,36.5,40.5,43.5),
                         y = c(0.5,4.5,8.5,11.5,13.5,15.5,20.5,23.5,28.5,32.5,36.5,40.5,43.5),
                         label = c("Size with other traits", "Water content",
                                   "Ash content", "Nitrogen content", 
                                   "Phosphorus content", "Proximate composition",
                                   "Chemical ratio","Energy content",
                                   "Respiration rate", "Excretion rate",
                                   "Feeding rate", "Growth", "Eggs"))


g1 <- ggplot() +
  
  # Polygons to highlight parts of the plot
  geom_polygon(data=green.box, aes(x=x, y=y), fill = "green", alpha = 0.2) +
  geom_polygon(data=yellow.box, aes(x=x, y=y, group = label), fill = "yellow", 
               alpha = 0.3) +
  geom_text(data=box.labels, aes(x=x, y=y, label = label), 
            hjust = 0, vjust = 1.1, size = 5) +
  
  
  # Correlation data
  geom_point(data = cor.table.2, aes(x = trait1.num, y = trait2.num, color = r.mag,
                        fill = r.phylo, size = N.species),
             shape = 21,) +
  geom_point(data = filter(cor.table.2, r.phylo >= 0.5),
             aes(x = trait1.num, y = trait2.num, fill = r.phylo, size = N.species),
             shape = 21, color = "black", stroke = 1) +
  # scale_color_gradient2(low = "blue3", mid = "white", high = "red2", 
  #                      limits = c(-1,1), name = "corr coef") +
  scale_color_manual(values = c("black",NA), guide = "none") +
  scale_fill_gradient2(low = "blue3", mid = "white", high = "red2", 
                       limits = c(-1,1), name = "r.phylo") +
  
  scale_x_discrete(limits = trait.list) +
  scale_y_reverse(breaks = 1:length(trait.list), labels = trait.list,
                  limits = c(length(trait.list)+0.5,0.5), expand = c(0, 0))+
  scale_size_continuous(breaks = c(10,100,200,250), name = "N") +
  # scale_y_discrete(limits = (trait.list))  +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,  hjust=1),
        axis.title = element_blank(),
        aspect.ratio = 1,
        legend.position = "bottom",
        # legend.box = "vertical",
        # legend.key.height = unit(0.5,"in"),
        legend.key.width=unit(0.3,"in"),
        text = element_text(size = 16),
        panel.grid.minor.y = element_blank(),
        panel.background = element_rect(fill='white'),
        plot.background = element_rect(fill='white', color=NA))

g1
```

# Categorical trait Cramer's V
## Prepare the data frame
Because some species can have multiple categories for a categorical trait, it is best to do this analysis with the binary versions. Only 8 traits have binary versions: feeding mode, reproduction mode, trophic group, vertical distribution, dvm, hibernation,and myelination but hibernation only includes positive records.
```{r}
# Filter the subset data to only traits which we have a good number of species to do correlations.
trait.binary <- traits.lvl2 %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(valueType == "binary") %>% 
  filter(!grepl("HIB.", traitName))

sort(unique(trait.binary$traitName))

FM <- trait.binary %>% 
  filter(grepl("FM.", traitName)) %>% 
  mutate(val = str_replace(traitName, "FM.","")) %>% 
  dplyr::select(taxonID, scientificName, val) %>% 
  mutate(traitName = "feedingMode")
FM1 <- FM %>% 
  filter(val %in% c("cruise","current","active.ambush","passive.ambush",
                           "particle.feeder"))
FM2 <- FM %>% 
  filter(val %in% c("active","passive"))

TG <- trait.binary %>% 
  filter(grepl("TG.", traitName)) %>% 
  mutate(val = str_replace(traitName, "TG.","")) %>% 
  dplyr::select(taxonID, scientificName, val) %>% 
  mutate(traitName = "trophicGroup")

RM <- trait.binary %>% 
  filter(grepl("RM.", traitName)) %>% 
  mutate(val = str_replace(traitName, "RM.","")) %>% 
  dplyr::select(taxonID, scientificName, val) %>% 
  mutate(traitName = "reproductionMode")

MYE <- trait.binary %>% 
  filter(grepl("MYE.", traitName)) %>% 
  filter(traitValue == 1) %>% 
  mutate(val = str_replace(traitName, "MYE.","")) %>% 
  dplyr::select(taxonID, scientificName, val)%>% 
  mutate(traitName = "myelination")

BIO <- trait.binary %>% 
  filter(grepl("BIO.", traitName)) %>% 
  filter(traitValue == 1) %>% 
  mutate(val = str_replace(traitName, "BIO.","")) %>% 
  dplyr::select(taxonID, scientificName, val) %>% 
  mutate(traitName = "bioluminescence")

VD <- traits.lvl2 %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(traitName == "verticalDistribution") %>% 
  dplyr::select(taxonID, scientificName, traitName, val = traitValue)

DVM <- traits.lvl2 %>% 
  filter(taxonRank %in% c("Species","Subspecies")) %>% 
  filter(traitName == "dielVerticalMigration") %>% 
  dplyr::select(taxonID, scientificName, traitName, val = traitValue) %>% 
  filter(val != "absent; present") %>% # Exclude Euphausia superba which has records for both
  mutate(val = if_else(grepl("absent",val), "absent", "present")) # only retain present or absent records

# Combine all to a single dataframe
trait.cat <- bind_rows(FM1, TG, RM, VD, DVM, MYE, BIO)
```


## Loop association test with Cramer's V
A measure of association between two categorical variables. It is a scaled version of the chi-squared test statistic and lies between 0 and 1. Note that there are no negative values, since statistic is bounded at [0,1], >0.5 is strong association.
```{r}
# Loop throguh pairs or rate traits, and calculate v
trait.list.cat <- c("feedingMode", "trophicGroup", "reproductionMode",
                "verticalDistribution", "dielVerticalMigration",
                "bioluminescence", "myelination")

trait.cor <- data.frame(trait1 = character(), trait2 = character())
for (i in c(1:(length(trait.list.cat)-1))) {
  for (j in c(i:length(trait.list.cat))) {
    trait.cor <- trait.cor %>% 
      bind_rows(data.frame(trait1 = trait.list.cat[i],
                           trait2 = trait.list.cat[j]))
  }
}
trait.cor <- trait.cor %>% 
  filter(trait1 != trait2)

v.results <- data.frame()
for(i in c(1:nrow(trait.cor))) {
  A <- trait.cat %>% 
    filter(traitName %in% trait.cor[i,1]) %>% 
    dplyr::select(taxonID, trait1 = val)
  B <- trait.cat %>% 
    filter(traitName %in% trait.cor[i,2]) %>% 
    dplyr::select(taxonID, trait2 = val)
  
  AB <- A %>% 
    left_join(B, by = "taxonID") %>% 
    filter(!is.na(trait2))
  
  AB.results <- data.frame(test.num = i,
                           trait1 = trait.cor[i,1],
                           trait2 = trait.cor[i,2],
                           N.pairs = nrow(AB),
                           N.species = length(unique(AB$taxonID)),
                           V = cramerV(AB$trait1, AB$trait2, 
                                       bias.correct = TRUE))
  v.results <- bind_rows(v.results, AB.results)
}
```

## Plot correlation matrix
Most pairs have weak association except for feeding mode and reproduction mode. There is weak association (0.3-0.5) between bioluminescence and myelination, DVM and vertical distribution, and feeding mode and trophic group.
```{r}
v.table <- v.results %>% 
  mutate(trait1 = factor(trait1, levels = trait.list.cat),
         trait2 = factor(trait2, levels = trait.list.cat))

v.table %>% 
  filter(abs(V) > 0.3) %>% 
  nrow()

g2 <- ggplot(v.table, aes(x = trait1, y = trait2,
                        fill = V, size = N.species)) +
  geom_point(shape = 21) +
  geom_point(data = filter(v.table, V >= 0.5),
             aes(x = trait1, y = trait2), shape = 21, color = "black", stroke = 1) +
  
  # scale_fill_distiller(palette = "YlOrRd", limits = c(0,1)) +
  scale_fill_gradient(low = "white", high = "red2",
                       limits = c(0,1), name = "V") +
  scale_size_continuous(breaks = c(100, 250, 500), name = "N") +
  scale_y_discrete(limits = rev(trait.list.cat)) +
  # scale_x_discrete(position = "top") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,  hjust=1),
        axis.title = element_blank(),
        text = element_text(size = 16),
        legend.justification = "top",
        # legend.position = "bottom",
        # legend.box = "vertical",
        # legend.key.height = unit(0.25,"in"),
        # legend.position = c(1.2, 0.75),
        # legend.box.just = "top",
        aspect.ratio = 1,
        # panel.background = element_rect(fill='white'),
        plot.background = element_rect(fill='white', color = "black")) 
g2
```


# Figure 6. Association between traits. 
Inset categorical on numerical trait plot
```{r, fig.height=5, fig.width=5}
cowplot::ggdraw() +
  draw_plot(g1) +
  draw_plot(g2, x = 0.61, y = .645, width = 0.325, height = 0.4)
  
ggsave("figures/fig_6_association_between_traits.png",
       width = 14, height = 13, units = "in", dpi = 600)
```
