---
title: "FW Trait Data Standardization"
author: "Patrick Pata"
date: "2025-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Hebert et al 2016 Zoop data - FW portion
2. Beisner Functional Trait Matrix
3. Obertegger et al 2023
4. Colina et al 2016
5. Barnett 2007_2013

Note: Later, need a WoRMS Taxonomy table for all taxa names

# Load packages and functions
```{r}
library(tidyverse)
library(here)
source("../toolkit.R")

# The base format of the trait database data frame.
s.format <- read.csv(here("data_input/trait_dataset_standard_format_20230628.csv"))[-1,]

# Trait directory for converting verbatim trait names
trait.directory <- read.csv(here("data_input/trait_directory_2025.csv")) %>% 
  select(-c(traitBucket, traitNum, trait))

getref <- function(refval = NA, refs.list = refs) {
  # If multiple vals seperated by comma, separate
  rfs <- as.numeric(unlist(str_split(refval, ",")))
  # Extract from the refs list
  ref.sub <- refs.list %>% 
    filter(Ref.code %in% rfs) 
  df <- paste0(ref.sub$ref_id, collapse = "; ")
  return(df)
}

```

# Hebert et al. 2016 - Freshwater
## Read data
- Note that Macrosetella sp. has no habitat type so excluded in previous curation
```{r}
# Original file
f <- here("dev_files/literature_data/Hebert2016/zooplankton_traits.csv")

data <- read.csv2(f, sep = ";", dec = ",") %>% 
  filter(Habitat == "Freshwater")

# reference table
refs <- read.csv(here("dev_files/literature_data/Hebert2016/references_ed_v2.csv"))


# Replace non numeric rows to numeric and , to .
data.class <- unlist( lapply(data, class) ) %>% 
  as.data.frame() %>% 
  rename(Class = ".") %>% 
  rownames_to_column("ColumnName") %>% 
  filter(Class == "character")

# The rest of the coercion warning are when the values are "RT" or "nd " (room temperature and not detectable)
data <- data %>% 
  # mutate(Replicate.number = as.numeric(Replicate.number)) %>% # Also References?
  mutate(Min.dm = as.numeric(str_replace(str_replace(Min.dm," ",""),",","."))) %>%
  mutate(Min.lip = as.numeric(str_replace(str_replace(Min.lip," ",""),",","."))) %>% 
  mutate(Temp.resp.rt = as.numeric(str_replace(str_replace(Temp.resp.rt," ",""),",","."))) %>%
  mutate(Min.Nex.rt.ug.dm = as.numeric(str_replace(str_replace(Min.Nex.rt.ug.dm," ",""),",","."))) %>% 
  mutate(Temp.Nex.rt = as.numeric(str_replace(str_replace(Temp.Nex.rt," ",""),",","."))) %>%
  mutate(P.phosphate.ex.rt.ug.ind = as.numeric(str_replace(str_replace(P.phosphate.ex.rt.ug.ind," ",""),",","."))) %>%
  mutate(Temp.Pex.rt = as.numeric(str_replace(str_replace(Temp.Pex.rt," ",""),",","."))) %>%
  mutate(Min.NP.ex = as.numeric(str_replace(str_replace(Min.NP.ex," ",""),",",".")))

# For each trait, expand to long format, separate column for references and add unit column
heb.tg <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Trophic.group, Ref.tg) %>% 
  filter(!is.na(Trophic.group)) %>% 
  rename(verbatimTraitValue = Trophic.group,
         observationNumber = Replicate.number) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         verbatimTraitName = "Trophic.group",
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.tg) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code)
  # TODO recode trait value?

heb.bl <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Body.length, Min.bl, Max.bl, Ref.bl) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Body.length, Min.bl, Max.bl),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.bl) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "mm")
  

heb.dm <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Dry.mass, Min.dm, Max.dm, Ref.dm) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Dry.mass, Min.dm, Max.dm),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.dm) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "mg")

heb.C <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Cct.prop, Min.Cprop, Max.Cprop, Cct.total, Ref.C) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Cct.prop, Min.Cprop, Max.Cprop, Cct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.C) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Cct.total",
                                     "mg", verbatimTraitUnit))

heb.N <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Nct.prop, Min.Nprop, Max.Nprop, Nct.total, Ref.N) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Nct.prop, Min.Nprop, Max.Nprop, Nct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.N) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Nct.total",
                                     "mg", verbatimTraitUnit))


heb.P <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Pct.prop, Min.Pprop, Max.Pprop, Pct.total, Ref.P) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Pct.prop, Min.Pprop, Max.Pprop, Pct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.P) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Pct.total",
                                     "mg", verbatimTraitUnit))

heb.NP <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           N.Pct, Min.NP, Max.NP, Ref.NP) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.Pct, Min.NP, Max.NP),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.NP) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) 

heb.prot <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Protein.ct, Min.prot, Max.prot, Ref.prot) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Protein.ct, Min.prot, Max.prot),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.prot) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "%")

heb.lip <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Lipid.ct, Min.lip, Max.lip, Ref.lip) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Lipid.ct, Min.lip, Max.lip),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.lip) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code) %>% 
  mutate(verbatimTraitUnit = "%")

# respiration ++

heb.resp <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Respiration.rt.ind, Respiration.rt.mg, Min.resp.mg, Max.resp.mg,
           Temp.resp.rt, Min.temp.resp.rt, Max.temp.resp.rt, Ref.resp) %>% 
  rename(observationNumber = Replicate.number) %>% 
  select(where(~!all(is.na(.x)))) %>%  # remove rows with all NAs
  filter(!is.na(Ref.resp)) %>% 
  rename(verbatimTraitValue = Respiration.rt.ind,
         verbatimTemperature = Temp.resp.rt) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         verbatimTraitName = "Respiration.rt.ind",
         verbatimTraitUnit = "µl O2 ind-1 h-1",
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = Ref.resp) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species, Ref.code)


heb.Nex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           N.ammonia.ex.rt.ug.ind:Ref.N.ex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.ammonia.ex.rt.ug.ind:Max.Nex.rt.nmol.ind),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  pivot_longer(cols = c(Temp.Nex.rt, Min.temp.Nex.rt, Max.temp.Nex.rt),
               names_to = "notes",
               values_to = "verbatimTemperature") %>% 
  filter(!is.na(verbatimTemperature)) %>% 
  # Verbatim temperature can be mid point of the min-max values
  group_by(Group, Genus, Species, observationNumber, Ref.N.ex,
           verbatimTraitName, verbatimTraitValue) %>% 
  mutate(notes = paste("Temperature variable based on ",
                       paste(notes, collapse = ","),
                       paste(as.character(verbatimTemperature), collapse = ","))) %>% 
  mutate(verbatimTemperature = mean(verbatimTemperature)) %>% 
  distinct() %>% 
  ungroup() %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016",
         verbatimTraitUnit = "") %>% 
  mutate(verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.ind") == TRUE,
                                     "µg N-NH4+ ind-1 h-1", verbatimTraitUnit),
         verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.dm") == TRUE,
                                     "µg N-NH4+ mgDM-1 h-1", verbatimTraitUnit)
         ) %>% 
  mutate(Ref.code = Ref.N.ex) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber, notes,
         Group, Genus, Species, Ref.code)



heb.Pex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           P.phosphate.ex.rt.ug.ind:Ref.P.ex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(P.phosphate.ex.rt.ug.ind:Max.Pex.rt.nmol.ind),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  pivot_longer(cols = c(Temp.Pex.rt, Min.temp.Pex.rt, Max.temp.Pex.rt),
               names_to = "notes",
               values_to = "verbatimTemperature") %>% 
  filter(!is.na(verbatimTemperature)) %>% 
  # Verbatim temperature can be mid point of the min-max values
  group_by(Group, Genus, Species, observationNumber, Ref.P.ex,
           verbatimTraitName, verbatimTraitValue) %>% 
  mutate(notes = paste("Temperature variable based on ",
                       paste(notes, collapse = ","),
                       paste(as.character(verbatimTemperature), collapse = ","))) %>% 
  mutate(verbatimTemperature = mean(verbatimTemperature)) %>% 
  distinct() %>% 
  ungroup() %>% 
  
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016",
         verbatimTraitUnit = "") %>% 
  mutate(verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.ind") == TRUE,
                                     "µg P-PO43- ind-1 h-1", verbatimTraitUnit),
         verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.dm") == TRUE,
                                     "µg P-PO43- mgDM-1 h-1", verbatimTraitUnit)
         ) %>% 
  mutate(Ref.code = Ref.P.ex) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber, notes,
         Group, Genus, Species, Ref.code)


heb.NPex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           N.P.ex:Ref.NPex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.P.ex, Min.NP.ex, Max.NP.ex),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = as.character(Ref.NPex)) %>% 
  group_by(Ref.code) %>% 
  mutate(primaryReference = getref(Ref.code)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber, 
         Group, Genus, Species, Ref.code)

# Add some of the marine NP excretion ratios not included in the original database version
heb.NPex.marine <- read.csv2(f, sep = ";", dec = ",") %>% 
  filter(Habitat == "Marine") %>% 
  mutate(Min.NP.ex = as.numeric(str_replace(str_replace(Min.NP.ex," ",""),",","."))) %>% 
  select(Group, Genus, Species, Replicate.number,
           N.P.ex:Ref.NPex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.P.ex, Min.NP.ex, Max.NP.ex),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  mutate(Ref.code = as.character(Ref.NPex)) %>% 
  group_by(Ref.NPex) %>% 
  mutate(primaryReference = getref(Ref.NPex)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber, 
         Group, Genus, Species, Ref.code)


# Determine the associate size value?
dm.sub <- heb.dm %>% 
  filter(verbatimTraitName == "Dry.mass") %>% 
  select(verbatimScientificName, observationNumber,
         sizeAssocName = verbatimTraitName,
         sizeAssocValue = verbatimTraitValue,
         sizeAssocUnit = verbatimTraitUnit,
         sizeAssocReference = primaryReference) %>% 
  mutate(sizeAssocName = "dryWeight")

# Combine all rows
heb.long <- bind_rows(heb.C, heb.N, heb.P, heb.NP,
                      heb.lip, heb.prot,
                      heb.resp, heb.Nex, heb.Pex, heb.NPex,
                      heb.NPex.marine) %>% 
  # Add size associated values for physiology and composition traits
  left_join(dm.sub,
            by = join_by(verbatimScientificName, observationNumber)) %>% 
  # Add the size traits
  bind_rows(heb.bl, heb.dm) %>% 

# Recode traitNames and units, add traitID and valueType
  mutate(verbatimTraitUnit = if_else(verbatimTraitUnit == "%",
                                     "percent", verbatimTraitUnit)) %>% 
  mutate(verbatimTraitUnit = if_else(is.na(verbatimTraitUnit),
                                     "", verbatimTraitUnit)) %>% 
  mutate(verbatimTraitUnit = str_replace(verbatimTraitUnit, "µ", "u")) %>%
  mutate(verbatimTraitUnit = str_replace(verbatimTraitUnit, "mgDM-1", "mg DM-1")) %>% 
  left_join(trait.directory, by = join_by(verbatimTraitName, verbatimTraitUnit)) %>% 
  mutate(traitValue = verbatimTraitValue,
    traitValueTemperature = verbatimTemperature)



# Recoding categorical traits
# NOTE that because heb.tg is categorical, separate it for now before unit conversions of the numerical traits
heb.long.tg <- heb.tg %>% 
  mutate(verbatimTraitUnit = "") %>% 
  left_join(trait.directory, by = join_by(verbatimTraitName, verbatimTraitUnit)) %>% 
  mutate(traitValue = tolower(verbatimTraitValue)) %>% 
  mutate(traitValue = if_else(verbatimTraitValue == "OmniHerb",
                              "omnivore; herbivore", traitValue),
         traitValue = if_else(verbatimTraitValue == "OmniCarn",
                            "omnivore; carnivore", traitValue),
         traitValue = if_else(verbatimTraitValue == "Predator",
                            "carnivore", traitValue)) 
 

# TODO find references not in master reference list and organize DOIs
# TODO rename some refs to match how it is coded in the master list (e.g. Ikeda1974 can be Ikeda1974a/b)
master.ref <- read.csv(here("data_input/reference_list_2025.csv"))
refs2organize.heb <- refs %>% 
  filter(ref_id %notin% master.ref$referenceCode)

# TODO merge and organize...convert numerics to character
colnames(heb.long)
colnames(heb.long.tg)

level0.heb <- heb.long %>% 
  mutate(traitValue = as.character(traitValue)) %>% 
  mutate(verbatimTraitValue = as.character(verbatimTraitValue)) %>% 
  # set life stage (assume default is female)
  mutate(lifeStage = if_else(grepl("(M)", Species, fixed = TRUE), "M","F")) %>% 
  mutate(stageID = if_else(grepl("(M)", Species, fixed = TRUE),3, 2)) %>% 
  bind_rows(heb.long.tg)


# Save the level0 data...
save(file = here("dev_files/standardized_datasets_dfs/level0_Hebert2016.RData"),
     level0.heb, refs2organize.heb)

rm(heb.C, heb.N, heb.P, heb.NP, heb.lip, heb.prot,
   heb.resp, heb.Nex, heb.Pex, heb.NPex,
   dm.sub,heb.bl, heb.dm, heb.NPex.marine, heb.tg,
   heb.long, heb.long.tg, refs)
```

## Standardize data from level0 to level1 standardized
Unit conversion of numerical traits not necessary because already follows the standard
```{r, eval=FALSE}
# update numeric data by removing all _Min or _Max values and appending the derived midpoint values
data.midpoint <- heb.long %>% 
  filter(grepl("_Min|_Max", traitName)) %>% 
  mutate(traitValue = as.numeric(verbatimTraitValue)) %>% 
  mutate(traitName  = gsub("\\_Min","",traitName),
         traitName = gsub("\\_Max","",traitName)) %>% 
  group_by(traitName, traitUnit, observationNumber,
           primaryReference, secondaryReference) %>% 
  mutate(traitValueN = n(),
         traitValue = mean(traitValue), 
         notes = cleanStrings(paste(notes,"; value is midpoint of min and max")),
         verbatimTraitName = paste0(verbatimTraitName, collapse = "; "),
         verbatimTraitValue = paste0(verbatimTraitValue, collapse = "; "),
         isDerived = TRUE,
         traitValueSource = "derived from midpoint of range") %>% 
  ungroup()  %>% 
  distinct(traitValue,
           traitName, traitUnit, 
           primaryReference, secondaryReference, .keep_all = TRUE) %>% 
  standardizeID(trait.directory) %>% 
  # NA observation numbers because these are new records and will be assigned new IDs
  mutate(traitTaxonID = NA, observationNumber = NA, maxObsNum = NA)
  
heb.long.2 <- heb.long %>% 
  filter(!(grepl("_Min|_Max", traitName))) %>% 
  bind_rows(data.midpoint) 


# Apply weight specific conversions
data.WSDW <- heb.long.2 %>% 
  filter(traitName %in% c("respirationRate", "excretionRateN", "excretionRateP")) %>% 
  filter(!is.na(sizeAssocValue)) %>% 
  group_by(verbatimScientificName, primaryReference) %>% 
  mutate(traitValue = traitValue / sizeAssocValue,
         traitName = paste0(traitName,"_WSDW"),
         traitUnit = str_replace(traitUnit,"h","mg DW^-1 h"),
         notes = cleanStrings(paste(notes,"; calculated weight specific value from associated size value")),
         isDerived = TRUE,
         traitValueSource = "derived from individual-specific value") %>% 
  ungroup() %>% 
  # reassign trait ID 
  standardizeID(trait.directory) %>% 
  mutate(traitTaxonID = NA, observationNumber = NA, maxObsNum = NA)

heb.long.2 <- bind_rows(heb.long.2, data.WSDW) 

# TODO Apply temperature standardization
data.rates.resp <- heb.long.2 %>% 
  filter(traitName %in% c("respirationRate",
                          "respirationRate_WSDW"))  %>% 
  mutate(traitValueTemperature = as.numeric(traitValueTemperature)) %>% 
  filter(!is.na(traitValueTemperature)) %>% 
  group_by(traitTaxonID) %>% 
  mutate(traitValue = standard_temp(traitValue, traitValueTemperature),
         traitName = paste0(traitName,"_15C"),
         traitValueTemperature = 15,
         notes = cleanStrings(paste(notes,"; value temperature standardized with Q10 = 2.8")),
         isDerived = TRUE,
         traitValueSource = "calculated using Q10 equation") %>% 
  ungroup()

# Excretion rates Ammonia Q10=2.0, phosphate Q10=1.55 (Based on Hebert et al. 2016)
data.rates.ex.N <- heb.long.2 %>% 
  filter(traitName %in% c("excretionRateN","excretionRateN_WSDW")) %>% 
  mutate(traitValueTemperature = as.numeric(traitValueTemperature)) %>% 
  filter(!is.na(traitValueTemperature)) %>% 
  group_by(traitTaxonID) %>% 
  mutate(traitValue = standard_temp(traitValue, traitValueTemperature,
                                    t = 15, Q10 = 2.0),
         traitName = paste0(traitName,"_15C"),
         traitValueTemperature = 15,
         notes = cleanStrings(paste(notes,"; value temperature standardized with Q10 = 2.0")),
         isDerived = TRUE,
         traitValueSource = "calculated using Q10 equation") %>% 
  ungroup()

data.rates.ex.P <- heb.long.2 %>% 
  filter(traitName %in% c("excretionRateP","excretionRateP_WSDW")) %>% 
  mutate(traitValueTemperature = as.numeric(traitValueTemperature)) %>% 
  filter(!is.na(traitValueTemperature)) %>% 
  group_by(traitTaxonID) %>% 
  mutate(traitValue = standard_temp(traitValue, traitValueTemperature,
                                    t = 15, Q10 = 1.55),
         traitName = paste0(traitName,"_15C"),
         traitValueTemperature = 15,
         notes = cleanStrings(paste(notes,"; value temperature standardized with Q10 = 1.55")),
         isDerived = TRUE,
         traitValueSource = "calculated using Q10 equation") %>% 
  ungroup()


# TODO save the dataset into a dataframe 


```



# Beisner trait matrix
*NEED TO CONFIRM PRIMARY REFERENCES*
- Also need to confirm if the sample column is just for specific study (For 8 species listed as both Sample = Pelagic and Sediment, the size value is the same, for 12 species, size is different)
- Note that some taxa names are "calanoid/cyclopoid copepodite"
```{r}
f <- here("dev_files/literature_data/Fresh water zooplankton traits - from B Beisner/Functionaltraitsmatrix.csv")

level0.beis <- read.csv(f) %>% 
# Change . in species name
  mutate(verbatimScientificName = Species.name) %>% 
  mutate(Species.name = str_trim(str_replace_all(Species.name,"\\."," "))) %>% 
# Convert to long format
  mutate(mean.length_mm = as.character(mean.length_mm)) %>% 
  pivot_longer(cols = c(mean.length_mm, feeding.type, habitat, trophic.group),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "mean.length_mm",
                                     "mm", ""),
         secondaryReference = "BeisnerData",
         # TODO find out references if present (maybe duplicated with the Barnett et al. 2007/Hebert et al. 2016)
         primaryReference = "TBD") %>% 
  # Remove instances when records are duplicated because there were separate rows for Sample = Pelagic/Sediment. For 12 species, the size trait is different between samples
  group_by(verbatimScientificName, verbatimTraitName, verbatimTraitValue) %>% 
  filter(row_number() == 1) %>% 
  mutate(notes = paste("Sample =", Sample)) %>% 
  ungroup() %>% 
  select(-Sample) %>% 
  # standardize traitnames and units
  left_join(trait.directory,  by = join_by(verbatimTraitName, verbatimTraitUnit)) %>% 
  mutate(traitValue = tolower(verbatimTraitValue)) %>% 
  # rename levels of categorical data
  mutate(traitValue = case_when(
    verbatimTraitValue == "Interm" ~ "intermediate",
    verbatimTraitValue == "Omnivore/Carnivore" ~ "omnivore; carnivore",
    verbatimTraitValue == "Omnivore/Herbivore" ~ "omnivore; herbivore",
    verbatimTraitValue == "current-cruiser" ~ "current; cruise",
    verbatimTraitValue == "Raptorial" ~ "cruise",
    .default =  traitValue))

save(file = here("dev_files/standardized_datasets_dfs/level0_BeisnerData.RData"),
     level0.beis)

```

# Colina et al. 2016
The appendix.csv and references.csv are the spreadsheet version of the supplementary word file. The references column was manually annotated to fix typos and change special characters.
- Note: clearance (CR, mL.ind-1 d-1) and ingestion rates (IR, ugC.ind-1.d-1), phytoplankton initial concentration (C0, mgC.L-1), MBFG affiliation of the phytoplankton, and references.
- Note: no associated temperature or size information 

MBFG	Description
I	Small organisms with high sruface/volume
II	Small-flagellated organisms with silicieous exoskeletal structures
III	Large filaments with aerotopes
IV	Organisms of medium size lacking specialized traits
V	Unicellular flagellates of medium to large size
VI	Non-flagellated organisms with siliceous exoskeletons
VII	Large mucilaginous colonies

```{r}
data <- read.csv(
  here("dev_files/literature_data/Colina 2016/Colina2016_supplementary_appendix.csv"))
refs2organize.coli <- read.csv(
  here("dev_files/literature_data/Colina 2016/Colina2016_references.csv"))
mbfg <- read.csv(
   here("dev_files/literature_data/Colina 2016/MBFG_table.csv"))

level0.coli <- data %>% 
# rename columns and assign units
  rename(verbatimScientificName = Zooplankton.species) %>% 
  pivot_longer(cols = c("CR..mL.ind.1.d.1.","IR..ugC.ind.1.d.1."),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  mutate(verbatimTraitUnit = if_else(
    verbatimTraitName == "CR..mL.ind.1.d.1.", 
    "mL ind-1 d-1", "ugC ind-1 d-1")) %>% 
  mutate(verbatimTraitName = str_trim(str_replace_all(verbatimTraitName,"\\."," "))) %>%
# reassign MBFG descriptions and phyto concentration to notes
  left_join(mbfg, by = "MBFG") %>% 
  mutate(notes = paste("Phytoplankton initial concentration C0 =",
                       C0..mgC.L.1., "mg C L-1; Prey type =", MBFG_descriptor)) %>% 
# reassign references
  left_join(refs2organize.coli %>% 
              select(primaryReference, Reference_ed = Reference),
            by = "Reference_ed") %>% 
  mutate(secondaryReference = "Colina2016") %>% 
  select(-c(Reference, Reference_ed, C0..mgC.L.1., MBFG, MBFG_descriptor)) %>% 
  # remove blanks in trait value
  mutate(verbatimTraitValue = as.character(verbatimTraitValue)) %>% 
  filter(!is.na(verbatimTraitValue)) %>% 

# TODO standardize traitNames and Units

  left_join(trait.directory,  by = join_by(verbatimTraitName, verbatimTraitUnit)) %>% 
  filter(verbatimTraitValue != 0) %>% 
  filter(verbatimTraitValue != "") %>% 
  mutate(verbatimTraitValue = str_replace(verbatimTraitValue, ",", ".")) %>% 
  mutate(traitValue = verbatimTraitValue) 
    
save(file = here("dev_files/standardized_datasets_dfs/level0_Colina2016.RData"),
     level0.coli, refs2organize.coli)
```

# Obertegger and Wallace 2023 Rotifers 
- Traits at genus level
- Feeding traits were from specific references except some feeding trait values that were filled in based on literature. These were only marked by color in the spreadsheet so these needed to be manually organized to long format.
- Note that for the column headers that cite references as sensu/after CITATION, these references are not in the provided list and it is unclear if the values assigned colored in black were based on these references or the reference was just used for the terminology. For now, consider these as the primary reference but need to check the individual sources later.
- Although sometimes there are references in the column "literature to fill missing information on feeding traits" but there are no associated trait value in the blue text, so these references are actually for the black text?
- Some values in blue text have no reference citated and these were marked by NA in the annotation
- Manually revise reference for Harringia trophic group: macrofilter-feeders; raptors is [35]
- No clear references for most other than habitat
- In TableS2 notes: "Corona defined by possessing cilia used in swimming &/or food capture; not as an infundibulum [20]"

TODO: Standardize the coding of the feeding traits:
There are 4 feeding related traits: 
 1. feeding mechanism: ciliary movement, active capture, infundibulum 
 2. feeding niche: predator, microphagous TAXA, parasite, polyphagus, endosymbiotic algae in gut
 3. feeding type: phytophage, zoophage, trypot-bacterio, parasite [fuzzy]
 4. size of ingested particles: macrophages, microphages, parasite
```{r}
data <- read.csv(
  here("dev_files/literature_data/Rotifers Obertegger and Wallace 2023/Obertegger2023_supplementarytable2_part1.csv"))
refs2organize.ober <- read.csv(
  here("dev_files/literature_data/Rotifers Obertegger and Wallace 2023/Obertegger2023_references.csv"))
data_feeding <- read.csv(
  here("dev_files/literature_data/Rotifers Obertegger and Wallace 2023/Obertegger2023_supplementarytable2_part2.csv"))


# Organize habitat
obertegger.1 <- data %>% 
  select(genus, family, habitat, habitatref, habitat.coding.summarizing.information) %>% 
  mutate(verbatimScientificName = genus,
         verbatimTraitValue = habitat.coding.summarizing.information,
         verbatimTraitName = "habitat coding summarizing information",
         notes = habitat, # This is an elaborated habitat trait value with the reference
         secondaryReference = "Obertegger2023") %>% 
  # extract the reference
  group_by(habitatref) %>% 
  mutate(primaryReference = getref(habitatref, refs.list = refs2organize.ober)) %>% 
  ungroup() %>% 
  select(-c(habitat, habitatref, habitat.coding.summarizing.information)) %>% 
  #  standardize traitName and trait value coding
  left_join(trait.directory) %>% 
  mutate(traitValue = case_when(
    verbatimTraitValue == "littoral, periphytic" ~ "littoral; periphytic",
    verbatimTraitValue == "planktonic or semi-planktonic" ~ "planktonic; semi-planktonic",
    verbatimTraitValue == "littoral or periphytic; sometimes in plankton" ~ 
      "littoral; periphytic",
    .default = verbatimTraitValue
  ))

# Organize Trophi to Corona
obertegger.2 <- data %>% 
  select(genus, family, trophi.type, lorica, defence, foot, corona, coronaRef) %>% 
  mutate(verbatimScientificName = genus,
         secondaryReference = "Obertegger2023") %>% 
  pivot_longer(cols = c(trophi.type, lorica, defence, foot, corona),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  # assign the corona primary reference
  group_by(coronaRef, verbatimTraitName) %>% 
  mutate(primaryReference = if_else(verbatimTraitName == "corona",
         getref(coronaRef, refs.list = refs2organize.ober), NA)) %>% 
    ungroup() %>% 
  select(-coronaRef) %>% 
  relocate(genus, family, verbatimScientificName,
           verbatimTraitName, verbatimTraitValue,
           primaryReference, secondaryReference) %>% 

  #  standardize traitName and trait value coding
  left_join(trait.directory) %>% 
  mutate(traitValue = tolower(verbatimTraitValue))


# Oraganize feeding traits 
obertegger.3 <- data_feeding %>% 
  mutate(verbatimScientificName = genus,
         secondaryReference = "Obertegger2023") %>% 
  # the references in this column are expanded in Ref1-4, use this instead if this is the reference for all columns actually.
  pivot_longer(cols = c(feeding.mechanism.sensu.Monakov..1976., 
                        feeding.niche.categories.sensu.Gilbert..2022.,
                        feeding.types.sensu.Monakov..1976.,
                        size.of.ingested.particles.after.Monakov..1976.),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  mutate(verbatimTraitName = str_trim(str_replace_all(verbatimTraitName,"\\."," "))) %>% 
  # Assign reference based on traitName
  mutate(refCode = if_else(verbatimTraitName == "feeding mechanism sensu Monakov 1976",
                           REF1.Monakov..1976., NA),
         refCode = if_else(verbatimTraitName == "feeding niche categories sensu Gilbert 2022",
                           REF2.Gilbert..2022., refCode),
         refCode = if_else(verbatimTraitName == "feeding types sensu Monakov 1976",
                           REF3.Monakov..1976., refCode),
         refCode = if_else(verbatimTraitName == "size of ingested particles after Monakov  1976",
                           REF4.Monakov..1976., refCode)) %>% 
  
  group_by(refCode) %>%
  mutate(primaryReference = getref(refCode, refs.list = refs2organize.ober)) %>%
  ungroup() %>% 

  # assign the sensu and literature to fill missing info to notes
  mutate(notes = if_else(nchar(primaryReference) > 1,
                         paste(verbatimTraitName, ";reference to fill missing info:", primaryReference),
                         verbatimTraitName)) %>% 
  select(-c(refCode, REF1.Monakov..1976., REF2.Gilbert..2022., 
            REF3.Monakov..1976., REF4.Monakov..1976., 
            literature.to.fill.missing.information.on.feeding.traits)) %>% 
  relocate(genus, family, verbatimScientificName,
           verbatimTraitName, verbatimTraitValue,
           primaryReference, secondaryReference, notes) %>% 
  
  filter(nchar(verbatimTraitValue) > 1) %>% 
  
  #  standardize traitName and trait value coding
  left_join(trait.directory) %>% 
  mutate(traitValue = case_when(
    verbatimTraitValue == "ciliary movement creates strong currents concentrating food" ~ "current",
    verbatimTraitValue == "active capture" ~ "cruise",
    verbatimTraitValue == "infundibulum for enclosing & ingesting food" ~ "infundibulum",
    verbatimTraitValue == "parasite; not defined" ~ "parasite",
    verbatimTraitValue == "predator" ~ "carnivore",
    verbatimTraitValue == "not defined sensu Gilbert; feeding via endosybiotic algae in the gut" ~ 
      "endosymbiotic algae",
    verbatimTraitValue == "Macrophagous algivores" ~ "herbivore",
    verbatimTraitValue == "zoophages" ~ "carnivore",
    verbatimTraitValue == "phytophages" ~ "herbivore",
    verbatimTraitValue == "Macrophagous omnivores/predators" ~ "omnivore; carnivore",
    verbatimTraitValue == "Polyphagous rotifers" ~ "omnivore",
    verbatimTraitValue == "bacterio-phytophages" ~ "herbivore; bacterivore",
    verbatimTraitValue == "phytophages-zoophages" ~ "herbivore; carnivore",
    verbatimTraitValue == "trypto-bacterio" ~ "bacterivore",
    verbatimTraitValue == "trypto-bacterio-phytophages" ~ "herbivore; bacterivore",
    verbatimTraitValue == "trypto-bacterio-phytophages, zoophages" ~ "herbivore; carnivore; bacterivore",
    verbatimTraitValue == "trypto-bacteriophages" ~ "bacterivore",
    verbatimTraitValue == "Microphagous rotifers" ~ "bacterivore",
    .default = tolower(verbatimTraitValue)
  ))

# TODO combine
level0.ober <- bind_rows(obertegger.1, obertegger.2, obertegger.3)

save(file =  here("dev_files/standardized_datasets_dfs/level0_Obertegger2023.RData"),
     level0.ober, refs2organize.ober)
```

# Barnett et al 2007 FW zooplankton
**Note that primary references** are not provided for now because the encoding of the references and notes are not yet complete. 
## Create long table for annotation
```{r, eval=FALSE}
f <- here("dev_files/literature_data/Barnett 2007 FW zooplankton/Barnett_20072013_tables_corrigendum_working.xlsx")

# Temporary data
table1 <- openxlsx::read.xlsx(f, sheet = "table 1", startRow = 2)
table2 <- openxlsx::read.xlsx(f, sheet = "table 2", startRow = 2)
table3 <- openxlsx::read.xlsx(f, sheet = "table 3", startRow = 2)
table4 <- openxlsx::read.xlsx(f, sheet = "table 4", startRow = 2)

# Simply isolate species and traits
barnett1 <- table1 %>% 
  select(-c(Length_ref, Length_notes)) %>% 
  pivot_longer(cols = c(`Mean.body.length.(mm)`, `Food.size.range.(um)`,
         `Max.clearance.rate.(uL.ind-1.h-1)`, Regression.CR.vs.length,
         `Mesh.size.(um)`, `Fastest.generation.time.(days)`, 
         `Max.r.life.table.(day-1)`, `Max.r.population.growth.experiments.(day-1)`,
         `C/P`, `C/N`), 
         names_to = "verbatimTraitName",
         values_to = "verbatimTraitValue", 
         values_transform = as.character) %>% 
  mutate(secondaryReference = "Barnett2007") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(table = "table 1")

barnett2 <- table2 %>% 
  pivot_longer(cols = c(`Mean.body.length.(mm)`, `Food.size.range.(um)`,
                        `Max.clearance.rate.(uL.ind-1.h-1)`, 
                        `Fastest.generation.time.(days)`,
                        `C/P`, `C/N`), 
         names_to = "verbatimTraitName",
         values_to = "verbatimTraitValue", 
         values_transform = as.character) %>% 
  mutate(secondaryReference = "Barnett2007") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(table = "table 2")

barnett3 <- table3 %>% 
  select(-ends_with("_ref")) %>% 
  pivot_longer(cols = c(`Habitat`:`Food.size`), 
         names_to = "verbatimTraitName",
         values_to = "verbatimTraitValue", 
         values_transform = as.character) %>% 
  mutate(secondaryReference = "Barnett2007") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(table = "table 3")


barnett4 <- table4 %>% 
  select(-ends_with("_ref")) %>% 
  select(-ends_with("Ref")) %>% 
  pivot_longer(cols = c(`Trophic.group`:`Selectivity`), 
         names_to = "verbatimTraitName",
         values_to = "verbatimTraitValue", 
         values_transform = as.character) %>% 
  mutate(secondaryReference = "Barnett2007") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(table = "table 4")

barnett.all <- bind_rows(barnett1, barnett2, barnett3, barnett4) %>% 
  rename(verbatimScientificName = Species) %>% 
  # separate commas into rows
  separate_longer_delim(col = verbatimTraitValue, delim = ",") %>% 
  arrange(table, verbatimTraitName) %>% 
  relocate(table)

# # Temporary export for manual annotation of references
# write.csv(barnett.all, file = here("dev_files/literature_data/Barnett 2007 FW zooplankton/Barnett_20072013_forannotation.csv"),
#         row.names = F)

rm(barnett1, barnett2, barnett3, barnett4, barnett.all,
   table1, table2, table3, table4)
```

## Read long table for annotation
```{r}
# References
f <- here("dev_files/literature_data/Barnett 2007 FW zooplankton/Barnett_20072013_tables_corrigendum_working.xlsx")
refs2organize.barn <- openxlsx::read.xlsx(f, sheet = "References", startRow = 1) %>% 
  rename(Ref.code = Ref, ref_id = refcode, fullCitation = full.citation)

level0.barn <- read.csv(here("dev_files/literature_data/Barnett 2007 FW zooplankton/Barnett_20072013_annotated.csv")) %>% 

# Remove columns
  select(-c(table, Species_author, Group.counterpart, Species_Author)) %>% 

  # Update reference codes
  group_by(referenceCode) %>% 
  mutate(primaryReference = getref(referenceCode, refs.list = refs2organize.barn)) %>% 
  ungroup() %>% 

# Standardize trait names and units


  left_join(trait.directory, by = join_by(verbatimTraitName, verbatimTraitUnit)) %>% 
  
  # Standardize trait values
   mutate(traitValue = case_when(
    verbatimTraitValue == "Omnivore carnivore" ~ "omnivore; carnivore",
    verbatimTraitValue == "Omnivore herbivore" ~ "omnivore; herbivore",
    verbatimTraitValue == "Current-cruiser" ~ "current; cruise",
    verbatimTraitValue == "Raptorial" ~ "cruise",
    .default = tolower(verbatimTraitValue)
   )) %>% 
  # clearance rate from ul to ml
  mutate(traitValue = if_else(traitName == "clearanceRate_Max",
                              as.character(as.numeric(traitValue)/1000), 
                              traitValue)) %>% 
  mutate(assocTemperature = verbatimTemperature)
  


save(file =  here("dev_files/standardized_datasets_dfs/level0_Barnett2007.RData"),
     level0.barn, refs2organize.barn)
```


# Combine all FW trait data
## Load files
```{r}
load(here("dev_files/standardized_datasets_dfs/level0_Hebert2016.RData"))
load(here("dev_files/standardized_datasets_dfs/level0_Colina2016.RData"))
load(here("dev_files/standardized_datasets_dfs/level0_BeisnerData.RData"))
load(here("dev_files/standardized_datasets_dfs/level0_Obertegger2023.RData"))
load(here("dev_files/standardized_datasets_dfs/level0_Barnett2007.RData"))

```

## Merge reference list
```{r}
refs.new <- bind_rows(
  refs2organize.barn %>%  
    # select(-Ref.code) %>% 
    mutate(secondaryReference = "Barnett2007"),
  refs2organize.coli %>% 
    select(ref_id = primaryReference, fullCitation) %>% 
    mutate(secondaryReference = "Colina2016"),
  refs2organize.heb %>% 
    mutate(fullCitation = paste(Authors, Year, Title, Journal.Book)) %>% 
    select(ref_id, fullCitation, Ref.code) %>% 
    mutate(secondaryReference = "Hebert2016"),
  refs2organize.ober %>% 
    filter(ref_id != "") %>% 
    select(ref_id, fullCitation = Sources, Ref.code) %>% 
    mutate(Ref.code = as.numeric(Ref.code)) %>% 
    mutate(secondaryReference = "Obertegger2023") ) %>% 
 
  arrange(ref_id, secondaryReference) %>% 
  group_by(ref_id) %>% 
  mutate(ref_id_count = n()) %>% 
  mutate(rownum = row_number()) %>% 
  ungroup() %>% 
  arrange(-ref_id_count) %>% 
  # Default is to assign with either a/b/c
  mutate(primaryReference = ref_id) %>% 
  mutate(ref_id = if_else(ref_id_count > 1,
                          paste0(ref_id,letters[rownum]),
                          ref_id)) %>% 

  # Correct the repeated references
  mutate(ref_id = if_else(ref_id == "DeMott1982c",
                          "DeMott1982a", ref_id)) %>% 
  mutate(ref_id = if_else(ref_id == "Porter1980b",
                          "Porter1980", ref_id)) %>% 
  mutate(ref_id = if_else(ref_id == "Porter1980a",
                          "Porter1980", ref_id))

# change the ref_ids in the combined table
level0.barn <- level0.barn %>% 
  group_by(referenceCode) %>% 
  mutate(primaryReference = getref(refval = referenceCode, 
                                    refs.list = refs.new %>% 
                                     filter(secondaryReference == "Barnett2007"))) %>% 
  ungroup()

# Simply switch the primary reference codes
level0.coli <- level0.coli %>% 
  left_join(refs.new %>% 
              filter(secondaryReference == "Colina2016") %>% 
              filter(ref_id_count > 1) %>% 
              select(primaryReference, ref_id),
            by = "primaryReference") %>% 
  mutate(primaryReference = if_else(!is.na(ref_id),
                                    ref_id, primaryReference)) %>% 
  filter(traitValue != "") %>% 
  select(-ref_id)
  

# # Correct the Hebert2016 references by changing the reference file itself
# refs.heb.orig <- read.csv(here("dev_files/literature_data/Hebert2016/references_ed.csv")) %>% 
#   left_join(refs.new %>% 
#               filter(secondaryReference == "Hebert2016") %>% 
#               select(Ref.code, ref_id_new = ref_id)) %>% 
#   mutate(ref_id = if_else(!is.na(ref_id_new),
#                           ref_id_new, ref_id)) %>% 
#   select(-ref_id_new)
# 
# write.csv(refs.heb.orig, row.names = F,
#           file = here("dev_files/literature_data/Hebert2016/references_ed_v2.csv"))
 

# TODO annotate later with the DOIs


# TODO make sure not references found in the master list
master.ref <- read.csv(here("data_input/reference_list_2025.csv"))
refs.new.FW <- refs.new %>% 
  filter(ref_id %notin% master.ref$referenceCode)
```

## Merge data frames
```{r}
level0.FW <- bind_rows(
  level0.heb, level0.barn, level0.beis,
  level0.coli, level0.ober
) %>% 
  select(-c(Group, Genus, Species, Ref.code, referenceCode,
            Species.name, genus, family, Genus))


colnames(level0.FW)
```


## Create taxonomy table
```{r}
# List species and note duplicates
taxonomy.FW <- level0.FW %>% 
  mutate(name = str_replace_all(verbatimScientificName,fixed("."), " ")) %>% 
  mutate(name = str_replace_all(name, "-", " ")) %>% 
  mutate(name = str_replace_all(name, "/", " ")) %>% 
  mutate(name = cleanScientificName(name)) %>% 
  # rename some names
  mutate(name = case_when(
    name == "Daphnia galeata mendotae" ~ "Daphnia galeata",
    name == "Bosmina longispina maritima" ~ "Bosmina longispina",
    name == "Diacyclops bicuspifatus thomas" ~ "Diacyclops bicuspidatus",
    name == "Bosmina liederi longirostris" ~ "Bosmina liederi",
    name == "Daphnia cucullataxgaleata" ~ "Daphnia cucullata",
    name == "Daphnia hyalinaxgaleata" ~ "Daphnia hyalina",
    name == "" ~ "",
    name == "" ~ "",
    name == "" ~ "",
    name == "" ~ "",
    .default =  name)) %>% 
  
  select(verbatimScientificName, name) %>% 
  group_by(name) %>% 
  mutate(nrecords = n()) %>% 
  ungroup() %>% 
  distinct(name, .keep_all = T)

# Check if in the taxonomy table already
taxonomy.2023 <- read.csv(here("data_input/taxonomy_table_20230628.csv"))

taxaList.tosearch <- taxonomy.FW %>% 
  filter(name %notin% taxonomy.2023$scientificName) %>% 
  arrange(-nrecords)

# TODO find taxonomy information
library(worrms)
library(tictoc)

atablenow <- wm_records_taxamatch(name = taxaList.tosearch$name[1])
atablenow <- plyr::ldply(atablenow, data.frame)
atablenow <- atablenow[-1,]

tic("sp aphia search")
for (sp in taxaList.tosearch$name) {
  x <- tryCatch(wm_records_taxamatch(name = sp,
                                     marine_only = FALSE),
                error = function(e)
                print(paste("No aphia search for",sp)))
  if (is.list(x) == TRUE) {
    x <- plyr::ldply(x, data.frame)
    x$name_search <- sp
    x <- x %>% 
      filter(kingdom == "Animalia")
    
    if (nrow(x) > 1){
      print(sp)
      print(x$AphiaID)
      x <- x[1,]
    }
    
    atablenow <- rbind(atablenow, x)
  }
}

atablenow <- atablenow %>% 
  rename(aphiaID = valid_AphiaID,
         scientificName = valid_name)
toc()

# Which still has no taxonomy info
no.taxonomy <- taxaList.tosearch %>% 
  filter(name %notin% atablenow$name)

# TODO Some scientific names are unaccepted or not in Worms at all, so no aphiaID

# Organize to make sure aphiaIDs are not duplicateds
AA <- atablenow %>% 
  select(-c(AphiaID:authority)) %>% 
  # select(colnames(taxonomy.2023))
  group_by(aphiaID) %>% 
  filter(n()>1)

# Check matching
BB <- atablenow %>% 
  # select(colnames(taxonomy.2023))
  group_by(aphiaID) %>% 
  filter(n()==1) %>% 
  relocate(name_search, scientificName, match_type ) %>% 
  arrange(match_type)
# TODO add the taxonomy info to level0.FW and prepare table to append to the main taxonomy table


```

## Export all merged files
```{r}
save(level0.FW,
     refs.new.FW,
     taxonomy.FW)

```



# --- Extra ---
## Old version: Merge all trait references into a single data frame
```{r}
fw.traits <- bind_rows(
  heb.long %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)),
  heb.long.tg,
  beisner.long %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)),
  colina.long %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)),
  obertegger.1 %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)), 
  obertegger.2 %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)), 
  obertegger.3 %>% mutate(verbatimTraitValue = as.character(verbatimTraitValue)),
  barnett.all
) %>% 
  # temporarily filter to core columns
  select(verbatimScientificName, 
         verbatimTraitName,
         verbatimTraitValue,
         secondaryReference,
         family, genus, Species.name, Group.counterpart) %>% 
    mutate(organizedSpeciesName = if_else(is.na(Species.name),
                                        verbatimScientificName,
                                        Species.name)) %>% 
  group_by(organizedSpeciesName) %>% 
  mutate(organizedGenus =  strsplit(organizedSpeciesName, " ")[[1]][1]) %>% 
  ungroup()


fw.traits %>% 
  group_by(secondaryReference) %>% 
  count()

nrecords <- fw.traits %>% 
  group_by(organizedSpeciesName) %>% 
  summarise(nrecords = n(), .groups = "drop")
ggplot(nrecords, aes(x = nrecords)) +
  geom_histogram() +
  theme_bw()

# TODO rgbif get taxonomy table

length(unique(fw.traits$verbatimScientificName))
length(unique(fw.traits$verbatimScientificName))

# # Create list of traitnames for standardization
# traits <- fw.traits %>% 
#   distinct(verbatimTraitName) %>% 
#   arrange(verbatimTraitName)
# # Export
# write.csv(traits, file = here("dev_files/traits_list_to_standardize.csv"))

# Load standardized trait names
traits.stand <- read.csv(here("dev_files/traits_list_standardized.csv"))
fw.traits <- fw.traits %>% 
  left_join(traits.stand, by = "verbatimTraitName")

```

# TODO: exploratory plots of trait coverage
30 trait names/categories
453 unstandardize species names
3560 taxa-trait records
1617 genus-trait records

the corona/food... are for rotifers
```{r}
# number of traits per genera
fw.summary <- fw.traits %>% 
  group_by(organizedGenus, standTraitName) %>% 
  summarise(nrecords = n(),
            .groups = "drop")

# number of genera with trait data
fws.1 <- fw.summary %>% 
  group_by(standTraitName) %>% 
  summarise(ngenera = n(),
            .groups = "drop") %>% 
  arrange(ngenera) %>% 
  mutate(standTraitName = factor(standTraitName,
                                 levels = unique(standTraitName)))

ggplot(fws.1, aes(x = standTraitName, y = ngenera)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw() +
  ylab("# Genera") + xlab("Trait")

ggsave(here("figures/FW_trait_genus_summary.png"),
            bg = "white", width = 6, height = 4, units = "in")
```

