---
title: "FW Trait Data Standardization"
author: "Patrick Pata"
date: "2025-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Hebert et al 2016 Zoop data - FW portion
2. Beisner Functional Trait Matrix
3. Obertegger et al 2023
4. Colina et al 2016
5. Barnett 2007_2013

Note: Later, need a WoRMS Taxonomy table for all taxa names

# Load packages and functions
```{r}
library(tidyverse)
source("../toolkit.R")

# The base format of the trait database data frame.
s.format <- read.csv(here("data_input/trait_dataset_standard_format_20230628.csv"))[-1,]


getref <- function(refval = NA, refs.list = refs) {
  # If multiple vals seperated by comma, separate
  rfs <- as.numeric(unlist(str_split(refval, ",")))
  # Extract from the refs list
  ref.sub <- refs.list %>% 
    filter(Ref.code %in% rfs) 
  df <- paste0(ref.sub$ref_id, collapse = "; ")
  return(df)
}
```

# Read data
## Hebert et al. 2016 - Freshwater
- Note that Macrosetella sp. has no habitat type so excluded in previous curation
```{r}
# Original file
f <- here("dev_files/literature_data/Hebert2016/zooplankton_traits.csv")

data <- read.csv2(f, sep = ";", dec = ",") %>% 
  filter(Habitat == "Freshwater")

# reference table
refs <- read.csv(here("dev_files/literature_data/Hebert2016/references_ed.csv"))


# Replace non numeric rows to numeric and , to .
data.class <- unlist( lapply(data, class) ) %>% 
  as.data.frame() %>% 
  rename(Class = ".") %>% 
  rownames_to_column("ColumnName") %>% 
  filter(Class == "character")

# The rest of the coercion warning are when the values are "RT" or "nd " (room temperature and not detectable)
data <- data %>% 
  # mutate(Replicate.number = as.numeric(Replicate.number)) %>% # Also References?
  mutate(Min.dm = as.numeric(str_replace(str_replace(Min.dm," ",""),",","."))) %>%
  mutate(Min.lip = as.numeric(str_replace(str_replace(Min.lip," ",""),",","."))) %>% 
  mutate(Temp.resp.rt = as.numeric(str_replace(str_replace(Temp.resp.rt," ",""),",","."))) %>%
  mutate(Min.Nex.rt.ug.dm = as.numeric(str_replace(str_replace(Min.Nex.rt.ug.dm," ",""),",","."))) %>% 
  mutate(Temp.Nex.rt = as.numeric(str_replace(str_replace(Temp.Nex.rt," ",""),",","."))) %>%
  mutate(P.phosphate.ex.rt.ug.ind = as.numeric(str_replace(str_replace(P.phosphate.ex.rt.ug.ind," ",""),",","."))) %>%
  mutate(Temp.Pex.rt = as.numeric(str_replace(str_replace(Temp.Pex.rt," ",""),",","."))) %>%
  mutate(Min.NP.ex = as.numeric(str_replace(str_replace(Min.NP.ex," ",""),",",".")))

# For each trait, expand to long format, separate column for references and add unit column
heb.tg <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Trophic.group, Ref.tg) %>% 
  filter(!is.na(Trophic.group)) %>% 
  rename(verbatimTraitValue = Trophic.group,
         observationNumber = Replicate.number) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         verbatimTraitName = "Trophic.group",
         secondaryReference = "Hebert2016") %>% 
    group_by(Ref.tg) %>% 
  mutate(primaryReference = getref(Ref.tg)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species)
  # TODO recode trait value?

heb.bl <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Body.length, Min.bl, Max.bl, Ref.bl) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Body.length, Min.bl, Max.bl),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.bl) %>% 
  mutate(primaryReference = getref(Ref.bl)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "mm")
  

heb.dm <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Dry.mass, Min.dm, Max.dm, Ref.dm) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Dry.mass, Min.dm, Max.dm),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.dm) %>% 
  mutate(primaryReference = getref(Ref.dm)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "mg")

heb.C <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Cct.prop, Min.Cprop, Max.Cprop, Cct.total, Ref.C) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Cct.prop, Min.Cprop, Max.Cprop, Cct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.C) %>% 
  mutate(primaryReference = getref(Ref.C)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Cct.total",
                                     "mg", verbatimTraitUnit))

heb.N <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Nct.prop, Min.Nprop, Max.Nprop, Nct.total, Ref.N) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Nct.prop, Min.Nprop, Max.Nprop, Nct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.N) %>% 
  mutate(primaryReference = getref(Ref.N)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Nct.total",
                                     "mg", verbatimTraitUnit))


heb.P <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Pct.prop, Min.Pprop, Max.Pprop, Pct.total, Ref.P) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Pct.prop, Min.Pprop, Max.Pprop, Pct.total),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.P) %>% 
  mutate(primaryReference = getref(Ref.P)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "%") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "Pct.total",
                                     "mg", verbatimTraitUnit))

heb.NP <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           N.Pct, Min.NP, Max.NP, Ref.NP) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.Pct, Min.NP, Max.NP),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.NP) %>% 
  mutate(primaryReference = getref(Ref.NP)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) 

heb.prot <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Protein.ct, Min.prot, Max.prot, Ref.prot) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Protein.ct, Min.prot, Max.prot),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.prot) %>% 
  mutate(primaryReference = getref(Ref.prot)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "%")

heb.lip <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Lipid.ct, Min.lip, Max.lip, Ref.lip) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(Lipid.ct, Min.lip, Max.lip),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.lip) %>% 
  mutate(primaryReference = getref(Ref.lip)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species) %>% 
  mutate(verbatimTraitUnit = "%")

# respiration ++

heb.resp <- data %>% 
  distinct(Group, Genus, Species, Replicate.number,
           Respiration.rt.ind, Respiration.rt.mg, Min.resp.mg, Max.resp.mg,
           Temp.resp.rt, Min.temp.resp.rt, Max.temp.resp.rt, Ref.resp) %>% 
  rename(observationNumber = Replicate.number) %>% 
  select(where(~!all(is.na(.x)))) %>%  # remove rows with all NAs
  filter(!is.na(Ref.resp)) %>% 
  rename(verbatimTraitValue = Respiration.rt.ind,
         verbatimTemperature = Temp.resp.rt) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         verbatimTraitName = "Respiration.rt.ind",
         verbatimTraitUnit = "µl O2 ind-1 h-1",
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.resp) %>% 
  mutate(primaryReference = getref(Ref.resp)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber,
         Group, Genus, Species)


heb.Nex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           N.ammonia.ex.rt.ug.ind:Ref.N.ex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.ammonia.ex.rt.ug.ind:Max.Nex.rt.nmol.ind),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  pivot_longer(cols = c(Temp.Nex.rt, Min.temp.Nex.rt, Max.temp.Nex.rt),
               names_to = "notes",
               values_to = "verbatimTemperature") %>% 
  filter(!is.na(verbatimTemperature)) %>% 
  # Verbatim temperature can be mid point of the min-max values?
  mutate(notes = paste("Temperature variable is", notes)) %>% 
  
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016",
         verbatimTraitUnit = "") %>% 
  mutate(verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.ind") == TRUE,
                                     "µg N-NH4+ ind-1 h-1", verbatimTraitUnit),
         verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.dm") == TRUE,
                                     "µg N-NH4+ mgDM-1 h-1", verbatimTraitUnit)
         ) %>% 
  group_by(Ref.N.ex) %>% 
  mutate(primaryReference = getref(Ref.N.ex)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber, notes,
         Group, Genus, Species)



heb.Pex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           P.phosphate.ex.rt.ug.ind:Ref.P.ex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(P.phosphate.ex.rt.ug.ind:Max.Pex.rt.nmol.ind),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  pivot_longer(cols = c(Temp.Pex.rt, Min.temp.Pex.rt, Max.temp.Pex.rt),
               names_to = "notes",
               values_to = "verbatimTemperature") %>% 
  filter(!is.na(verbatimTemperature)) %>% 
  # Verbatim temperature can be mid point of the min-max values?
  mutate(notes = paste("Temperature variable is", notes)) %>% 
  
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016",
         verbatimTraitUnit = "") %>% 
  mutate(verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.ind") == TRUE,
                                     "µg P-PO43- ind-1 h-1", verbatimTraitUnit),
         verbatimTraitUnit = if_else(str_detect(verbatimTraitName, "ug.dm") == TRUE,
                                     "µg P-PO43- mgDM-1 h-1", verbatimTraitUnit)
         ) %>% 
  group_by(Ref.P.ex) %>% 
  mutate(primaryReference = getref(Ref.P.ex)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         verbatimTemperature, verbatimTraitUnit,
         secondaryReference, primaryReference, observationNumber, notes,
         Group, Genus, Species)


heb.NPex <- data %>% 
  select(Group, Genus, Species, Replicate.number,
           N.P.ex:Ref.NPex) %>% 
  rename(observationNumber = Replicate.number) %>% 
  pivot_longer(cols = c(N.P.ex, Min.NP.ex, Max.NP.ex),
           names_to = "verbatimTraitName",
           values_to = "verbatimTraitValue") %>% 
  filter(!is.na(verbatimTraitValue)) %>% 
  mutate(verbatimScientificName = paste(Genus, Species),
         secondaryReference = "Hebert2016") %>% 
  group_by(Ref.NPex) %>% 
  mutate(primaryReference = getref(Ref.NPex)) %>% 
  ungroup() %>% 
  select(verbatimScientificName, verbatimTraitName, verbatimTraitValue,
         secondaryReference, primaryReference, observationNumber, 
         Group, Genus, Species)


# Determine the associate size value?
dm.sub <- heb.dm %>% 
  select(verbatimScientificName, observationNumber,
         sizeAssocName = verbatimTraitName,
         sizeAssocValue = verbatimTraitValue,
         sizeAssocUnit = verbatimTraitUnit,
         sizeAssocReference = primaryReference)

# TODO calculate the mid point fo size to prevent many-to-many left joins in Pex
# TODO maybe do this for the combined data frame so that elemental composition have values too
# Combine all rows
heb.long <- bind_rows(heb.C, heb.N, heb.P, heb.NP,
                      heb.lip, heb.prot,
                      heb.resp, heb.Nex, heb.Pex, heb.NPex) %>% 
  # Add size associated values for physiology and composition traits
  left_join(dm.sub,
            by = join_by(verbatimScientificName, observationNumber)) %>% 
  # Add the size traits
  bind_rows(heb.bl, heb.dm)

# NOTE that because heb.tg is categorical, separate if for now before unit conversions of the numerical traits
heb.long.tg <- heb.tg

rm(heb.C, heb.N, heb.P, heb.NP, heb.lip, heb.prot,
   heb.resp, heb.Nex, heb.Pex, heb.NPex,
   dm.sub,heb.bl, heb.dm, )
```

## Beisner trait matrix
*NEED TO CONFIRM PRIMARY REFERENCES*
- Also need to confirm if the sample column is just for specific study (For 8 species listed as both Sample = Pelagic and Sediment, the size value is the same, for 12 species, size is different)
- Note that some taxa names are "calanoid/cyclopoid copepodite"
```{r}
f <- here("dev_files/literature_data/Fresh water zooplankton traits - from B Beisner/Functionaltraitsmatrix.csv")

beisner.long <- read.csv(f) %>% 
# Change . in species name
  mutate(verbatimScientificName = Species.name) %>% 
  mutate(Species.name = str_trim(str_replace_all(Species.name,"\\."," "))) %>% 
# Convert to long format
  mutate(mean.length_mm = as.character(mean.length_mm)) %>% 
  pivot_longer(cols = c(mean.length_mm, feeding.type, habitat, trophic.group),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  mutate(verbatimTraitUnit = if_else(verbatimTraitName == "mean.length_mm",
                                     "mm", NA),
         secondaryReference = "BeisnerData") %>% 
  # Remove instances when records are duplicated because there were sepearate rows for Sample = Pelagic/Sediment. For 12 species, the size trait is different between samples
  group_by(verbatimScientificName, verbatimTraitName, verbatimTraitValue) %>% 
  filter(row_number() == 1) %>% 
  mutate(notes = paste("Sample =", Sample)) %>% 
  select(-Sample)
```

## Colina et al. 2016
The appendix.csv and references.csv are the spreadsheet version of the supplementary word file. The references column was manually annotated to fix typos and change special characters.
- Note: clearance (CR, mL.ind-1 d-1) and ingestion rates (IR, ugC.ind-1.d-1), phytoplankton initial concentration (C0, mgC.L-1), MBFG affiliation of the phytoplankton, and references.

MBFG	Description
I	Small organisms with high sruface/volume
II	Small-flagellated organisms with silicieous exoskeletal structures
III	Large filaments with aerotopes
IV	Organisms of medium size lacking specialized traits
V	Unicellular flagellates of medium to large size
VI	Non-flagellated organisms with siliceous exoskeletons
VII	Large mucilaginous colonies

```{r}
data <- read.csv(
  here("dev_files/literature_data/Colina 2016/Colina2016_supplementary_appendix.csv"))
refs <- read.csv(
  here("dev_files/literature_data/Colina 2016/Colina2016_references.csv"))
mbfg <- read.csv(
   here("dev_files/literature_data/Colina 2016/MBFG_table.csv"))

colina.long <- data %>% 
# rename columns and assign units
  rename(verbatimScientificName = Zooplankton.species) %>% 
  pivot_longer(cols = c("CR..mL.ind.1.d.1.","IR..ugC.ind.1.d.1."),
               names_to = "verbatimTraitName",
               values_to = "verbatimTraitValue") %>% 
  mutate(verbatimTraitUnit = if_else(
    verbatimTraitName == "CR..mL.ind.1.d.1.", 
    "mL ind-1 d-1", "ugC ind-1 d-1")) %>% 
# reassign MBFG descriptions and phyto concentration to notes
  left_join(mbfg, by = "MBFG") %>% 
  mutate(notes = paste("Phytoplankton initial concentration C0 =",
                       C0..mgC.L.1., "mg C L-1; Prey type =", MBFG_descriptor)) %>% 
# reassign references
  left_join(refs %>% select(primaryReference, Reference_ed = Reference),
            by = "Reference_ed") %>% 
  mutate(secondaryReference = "Colina2016") %>% 
  select(-c(Reference, Reference_ed, C0..mgC.L.1., MBFG, MBFG_descriptor)) %>% 
  # remove blanks in trait value
  mutate(verbatimTraitValue = as.numeric(verbatimTraitValue)) %>% 
  filter(!is.na(verbatimTraitValue))

```

TODO:
## Obertegger and Wallace 2023 Rotifers 
```{r}

```

## Barnett et al 2007 FW zooplankton
```{r}

```


